---
title: "Parkinson Meta-Analysis"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  pMD_data_type: "pathabundance_unstratified"
  overwrite_output: "yes"
  taxonomy_type: "GTDB"
---

```{r setup}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)

library(mia) # Bioconductor
library(tidyverse) # CRAN
library(magrittr) # CRAN
library(DT) # CRAN

# load or install install github libraries
library(parkinsonsMetagenomicData) # remotes::install_github("ASAP-MAC/parkinsonsMetagenomicData")
library(biobakeryUtils) # remotes::install_github("g-antonello/biobakeryUtils")
library(ChocoPhlAnTaxonomies) # remotes::install_github("g-antonello/ChocoPhlAnTaxonomies")
```

# Download microbiome data

## PD studies processed with the nextflowMAC pipeline

```{r}

processed_datasets <- read_tsv("https://raw.githubusercontent.com/ASAP-MAC/parkinsons_data_search/refs/heads/main/shotgun_samples/processed_datasets.tsv")

processed_datasets[,1:5] %>% 
  relocate(`doi.org link`, .after = `Sequence Acc. code`) %>% 
  datatable(
    caption = paste("List of studies processed with the aim to include them in 
                    the meta-analysis so far.",
    "Analysis date:", 
    Sys.Date()))

country_per_study.chr <- c(BedarfJR_2017 = "Germany", BoktorJC_2023 = "USA", BoktorJC_2023 = "USA", 
DuruIC_2024 = "Finland", JoS_2022 = "South Korea", LeeEJ_2024 = "South Korea", 
MaoL_2021 = "China", NishiwakiH_2024 = "Japan", QianY_2020 = "China", 
WallenZD_2022 = "Deep South USA", ZhangM_2023 = "China", NGRC = "USA", 
UAB = "Deep South USA")
```

## Load and filter metadata of `parkinsonsMetagenomicData`

```{r}
data("sampleMetadata", package = "parkinsonsMetagenomicData")

metadata_datasets_available <- filter(sampleMetadata, 
                                      # keep only studies that have human host
                                      host_species == "Homo sapiens",
                                      # keep only studies with Parkinson disease as target condition
                                      target_condition == "Parkinson Disease",
                                      # remove blanks from BoktorJC_2023
                                      !grepl("BLANK", sample_id)
                                      )

metadata_datasets_available[, sapply(metadata_datasets_available, function(x) sum(is.na(x))/length(x)) < 0.1] %>% 
  datatable()
```

## Available downloadable data types

```{r}
datatable(parkinsonsMetagenomicData::biobakery_files())
```

## Download parkinsonsMetagenomicData for `r params$pMD_data_type`

```{r, eval=params$pMD_data_type != "genefamilies_unstratified"}
# return the raw data as they come from pMD

raw.tse <- returnSamples(data_type = params$pMD_data_type,
                                            sample_data = metadata_datasets_available)

# enhance data by adding basic assays and restructuring rowData
parkinsonsMetaAnalysisData.tse <- pMD_enhance(input.tse = raw.tse, data_type = params$pMD_data_type)
```

## Replace taxonomy with GTDB if `data_type == "relative_abundance"` and `taxonomy_type == "GTDB"`

```{r, eval=(params$pMD_data_type == "relative_abundance" & params$taxonomy_type == "GTDB")}
parkinsonsMetaAnalysisData.tse <- ChocoPhlAnTaxonomies::tse_replace_mpa_with_GTDB_taxonomy(data.tse = parkinsonsMetaAnalysisData.tse, chocophlan_datestamp = "202403")
```

## Add curated information about the Country of the participants/samples

```{r add country of study participants manually}

parkinsonsMetaAnalysisData.tse@colData$Country <- country_per_study.chr[parkinsonsMetaAnalysisData.tse@colData$study_name]

table(parkinsonsMetaAnalysisData.tse@colData$Country, useNA = "ifany")
```

## Split Boktor JC into 2 studies according to the BioProject ID

```{r}
tmp <- parkinsonsMetaAnalysisData.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  mutate(
    study_name = gsub("\\.1$", "", paste(study_name, as.integer(as.factor(BioProject)), sep = "."))
  ) %>% 
  ungroup()

# Re-fix rownames
rownames(tmp) <- tmp$uuid

parkinsonsMetaAnalysisData.tse@colData <- DataFrame(tmp)
```

## Sanity checks before saving data

Before saving the TSE object, we need to make sure of the following:

  - Remove samples with all-0 abundances (any assay)
  - Remove samples that have even just 1 NA
  - Have all abundances assigned to the `UN*` group (UNCLASSIFIED, UNMAPPED,
  UNINTEGRATED)
  
### Remove samples with all zeros OR with at least 1 `NA` in any of the assays

```{r}
# look for any assay whether a sample has all-zero or has NAs
all_0.logical <- sapply(assays(parkinsonsMetaAnalysisData.tse), function(x){
  x2_0 <- colSums(x) == 0
  x2_NA <- colSums(is.na(x))
  
  return(x2_0 | x2_NA > 0)
})

# create a _any_ selection vector
samples_to_remove <- apply(all_0.logical, 1, any)

# remove samples
parkinsonsMetaAnalysisData.tse <- parkinsonsMetaAnalysisData.tse[, !samples_to_remove]

dim(parkinsonsMetaAnalysisData.tse)
```

### Samples with all counts/abundances assigned to `UNCLASSIFIED|UNMAPPED|UNINTEGRATED`

```{r}
all_abunds_UN <- sapply(assays(parkinsonsMetaAnalysisData.tse), function(x){
  x2 <- colSums(x[!grepl("UN", rownames(x)),, drop = FALSE]) == 0
  return(x2)
})

samples_to_remove <- apply(all_abunds_UN, 1, any)

# remove samples
parkinsonsMetaAnalysisData.tse <- parkinsonsMetaAnalysisData.tse[, !samples_to_remove]
dim(parkinsonsMetaAnalysisData.tse)
```


### prune all-0 taxa after sample filtering

```{r}
all_0_assay <- sapply(assays(parkinsonsMetaAnalysisData.tse), function(x){
  x2 <- rowSums(x) == 0
  return(x2)
})

features_to_remove <- apply(all_0_assay, 1, any)

parkinsonsMetaAnalysisData_pruned.tse <- parkinsonsMetaAnalysisData.tse[!features_to_remove, ]
dim(parkinsonsMetaAnalysisData_pruned.tse)
```

## Write TSE file

```{r, message=TRUE}
nameSuffix <- paste(parkinsonsMetagenomicData::biobakery_files()$tool[parkinsonsMetagenomicData::biobakery_files()$data_type == params$pMD_data_type], params$pMD_data_type, sep = "_")

fullOutName <- sprintf("parkinsonsMetagenomicData/%s.tse", nameSuffix)

dir.create(fullOutName, showWarnings = FALSE)

if(params$overwrite_output == "yes"){
  unlink(fullOutName, recursive = TRUE)
}

write_TSE_to_dir(parkinsonsMetaAnalysisData_pruned.tse, out.dir = fullOutName)

message(sprintf("Data saved as: %s", fullOutName))
```

## Check if just-written TSE is identical to the one in memory

It is basically identical, but `data.table::fwrite` has a precision of `1e-15`.
(see (#3862)[https://github.com/Rdatatable/data.table/issues/3862]).
In practice this precision is enough for bioinformatic analyses.

```{r}
re_read_tse <- read_TSE_from_dir(fullOutName)

all.equal(parkinsonsMetaAnalysisData_pruned.tse, re_read_tse, tolerance = 1e-16)
```

# Session Info

```{r}
sessionInfo()
```
