---
title: "Download public data from parkinsonsMetagenomicData"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_pipeline: "MetaPhlAn"
  overwrite_TSE_out: "yes"
---

```{r setup}
if(!(params$biobakery_pipeline %in% c("MetaPhlAn", "HUMAnN"))){
  stop("Biobakery pipeline requested not existent, choose between 'MetaPhlAn' and 'HUMAnN'")
}

library(knitr) # CRAN
library(mia) # Bioconductor
library(tidyverse) # CRAN
library(magrittr) # CRAN
library(DT) # CRAN
library(jsonlite) # CRAN

# load or install install github libraries
library(parkinsonsMetagenomicData) # remotes::install_github("ASAP-MAC/parkinsonsMetagenomicData")
library(biobakeryUtils) # remotes::install_github("g-antonello/biobakeryUtils")
library(ChocoPhlAnTaxonomies) # remotes::install_github("g-antonello/ChocoPhlAnTaxonomies")


opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)

output_data.path <- "~/parkinsonMetaAnalysis_data/01-pMD_datasets/"
```

# Introduction and caveats

There are many data types available through the `parkinsonsMetagenomicData` R
package (see **Available downloadable data types** for full list). HUMAnN,
however, provides a so-called "stratified" alternative to the data which could
be useful to see which taxa contribute to each data type (gene family, pathway).

```{r}
datatable(parkinsonsMetagenomicData::biobakery_files())
```

Stratified datasets and unstratified gene families are really large and 
downloading should be done differently.
See [working-with-large-parquet-files.Rmd](https://github.com/ASAP-MAC/parkinsonsMetagenomicData/blob/devel/vignettes/working-with-large-parquet-files.Rmd).
Also, bear in mind that reading a parquet file locally is still memory intensive,
consider having at least 64 GB of memory available.

This is why this script is only able to download data from MetaPhlAn and HUMAnN
unstratified pathways.

```{r}
mapping_names.df <- data.frame(
  data_type = c("relative_abundance", "pathabundance_unstratified"),
  biobakery_pipeline = c("MetaPhlAn", "HUMAnN"),
  output_name = c("relative_abundance", "pathways")  
)

pipeline <- params$biobakery_pipeline
output_name <- mapping_names.df$output_name[mapping_names.df$biobakery_pipeline == pipeline]
data_type <- mapping_names.df$data_type[mapping_names.df$biobakery_pipeline == pipeline]
```

# Download raw data from `parkinsonsMetagenomicData`

```{r}
sample_filtering_steps <- list()
```

## Filter samples that are usable

  1. From human stool metagenomes
  2. Study with focus on Parkinson disease
  3. Exclude blanks (ad-hoc case for BoktorJC_2023)

```{r}
data("sampleMetadata", package = "parkinsonsMetagenomicData")

sample_filtering_steps[["Unfiltered"]] <- sampleMetadata$uuid

metadata_datasets_available <- filter(sampleMetadata, 
                                      # keep only studies that have human host
                                      host_species == "Homo sapiens")
sample_filtering_steps[["Human stool metagenome"]] <- metadata_datasets_available$uuid

metadata_datasets_available <- filter(metadata_datasets_available, 
                                     # keep samples from studies that target parkinson disease
                                      target_condition == "Parkinson Disease")
sample_filtering_steps[["Study about Parkinson"]] <- metadata_datasets_available$uuid
metadata_datasets_available <- filter(metadata_datasets_available, 
                                      # remove blanks from BoktorJC_2023
                                      !grepl("BLANK", sample_id))

sample_filtering_steps[["Not Blanks"]] <- metadata_datasets_available$uuid

metadata_datasets_available[, sapply(metadata_datasets_available, function(x) sum(is.na(x))/length(x)) < 0.1] %>% 
  datatable()
```

## Download data for `r paste(pipeline, data_type)`

```{r}
# get the raw data as they come from pMD
parkinsonsMetaAnalysisData.tse <- returnSamples(data_type = data_type,
                                            sample_data = metadata_datasets_available)

sample_filtering_steps[["Retrieved from pMD"]] <- parkinsonsMetaAnalysisData.tse@colData$uuid
```

## Split Boktor JC into 2 studies according to the BioProject ID

```{r}
tmp <- parkinsonsMetaAnalysisData.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  mutate(
    study_name = gsub("\\.1$", "", paste(study_name, as.integer(as.factor(BioProject)), sep = "."))
  ) %>% 
  ungroup()

# Re-fix rownames
rownames(tmp) <- tmp$uuid

parkinsonsMetaAnalysisData.tse@colData <- DataFrame(tmp)
```

## Sanity checks before saving data

Before saving the TSE object, we need to make sure of the following:

  - Remove samples with all-0 abundances (any assay)
  - Remove samples that have even just 1 NA
  - Have all abundances assigned to the `UN*` group (UNCLASSIFIED, UNMAPPED,
  UNINTEGRATED)
  
### Remove samples with all zeros OR with at least 1 `NA` in any of the assays

```{r}
# look for any assay whether a sample has all-zero or has NAs
all_0.logical <- sapply(assays(parkinsonsMetaAnalysisData.tse), function(x){
  x2_0 <- colSums(x) == 0
  x2_NA <- colSums(is.na(x))
  
  return(x2_0 | x2_NA > 0)
})

# create a _any_ selection vector
samples_to_remove <- apply(all_0.logical, 1, any)

# remove samples
parkinsonsMetaAnalysisData.tse <- parkinsonsMetaAnalysisData.tse[, !samples_to_remove]

sample_filtering_steps[["Sample abundances not entirely zero"]] <- parkinsonsMetaAnalysisData.tse@colData$uuid

dim(parkinsonsMetaAnalysisData.tse)
```

### Samples with all counts/abundances assigned to `UNCLASSIFIED|UNMAPPED|UNINTEGRATED`

```{r}
all_abunds_UN <- sapply(assays(parkinsonsMetaAnalysisData.tse), function(x){
  x2 <- colSums(x[!grepl("UN", rownames(x)),, drop = FALSE]) == 0
  return(x2)
})

samples_to_remove <- apply(all_abunds_UN, 1, any)

# remove samples
parkinsonsMetaAnalysisData.tse <- parkinsonsMetaAnalysisData.tse[, !samples_to_remove]

sample_filtering_steps[["Sample abundances not entirely UNCLASSIFIED|UNMAPPED|UNINTEGRATED"]] <- parkinsonsMetaAnalysisData.tse@colData$uuid
dim(parkinsonsMetaAnalysisData.tse)
```

### prune all-0 features after filtering

```{r}
all_0_assay <- sapply(assays(parkinsonsMetaAnalysisData.tse), function(x){
  x2 <- rowSums(x) == 0
  return(x2)
})

features_to_remove <- apply(all_0_assay, 1, any)

parkinsonsMetaAnalysisData_pruned.tse <- parkinsonsMetaAnalysisData.tse[!features_to_remove, ]
dim(parkinsonsMetaAnalysisData_pruned.tse)
```

## colData curation

```{r}
# convert age to numeric
parkinsonsMetaAnalysisData_pruned.tse@colData$age <- as.numeric(parkinsonsMetaAnalysisData_pruned.tse@colData$age)

# Binary Parkinson variable (controls may carry other diseases)
parkinsonsMetaAnalysisData_pruned.tse@colData$parkinson_for_analysis <- factor(ifelse(grepl("Parkinson", parkinsonsMetaAnalysisData_pruned.tse@colData$disease), 1, 0))

# Recode missing age_group
parkinsonsMetaAnalysisData_pruned.tse@colData$age_group <- factor(ifelse(
  is.na(parkinsonsMetaAnalysisData_pruned.tse@colData$age_group),
  "Unknown",
  parkinsonsMetaAnalysisData_pruned.tse@colData$age_group
))

# Number of diseases (including PD)
parkinsonsMetaAnalysisData_pruned.tse@colData$number_diseases <- ifelse(
  parkinsonsMetaAnalysisData_pruned.tse@colData$disease == "Healthy",
  0,
  str_count(parkinsonsMetaAnalysisData_pruned.tse@colData$disease, ";") + 1
)

# Country conversion per study
country_per_study.chr <- c(
  BedarfJR_2017   = "Germany",
  BoktorJC_2023   = "USA",
  BoktorJC_2023.2 = "USA",
  DuruIC_2024     = "Finland",
  JoS_2022        = "South Korea",
  LeeEJ_2024      = "South Korea",
  MaoL_2021       = "China",
  NishiwakiH_2024 = "Japan",
  QianY_2020      = "China",
  WallenZD_2022   = "Deep South USA",
  ZhangM_2023     = "China",
  NGRC            = "USA",
  UAB             = "Deep South USA"
)

parkinsonsMetaAnalysisData_pruned.tse@colData$Country <- country_per_study.chr[parkinsonsMetaAnalysisData_pruned.tse@colData$study_name]
```

## Pipeline specific adjustments

```{r}
parkinsonsMetaAnalysisData_pruned.tse <- pMD_enhance(parkinsonsMetaAnalysisData_pruned.tse, data_type = data_type)

if(pipeline == "MetaPhlAn"){
  # reads_processed as numeric values  
  parkinsonsMetaAnalysisData_pruned.tse@colData$reads_processed <- as.numeric(gsub("#([0-9]+).*", 
    "\\1", parkinsonsMetaAnalysisData_pruned.tse@colData$reads_processed))
  
  # reorder and rename assays to counts; relative_abundance; pathways
  assays(parkinsonsMetaAnalysisData_pruned.tse) <- assays(parkinsonsMetaAnalysisData_pruned.tse)[c(3, 1, 2)]
}

if(pipeline == "HUMAnN"){
  # assume some reads processed from the abundances of the pathways
  parkinsonsMetaAnalysisData_pruned.tse@colData$reads_processed <- colSums(assay(parkinsonsMetaAnalysisData_pruned.tse))
  
  # rename first assay
  assayNames(parkinsonsMetaAnalysisData_pruned.tse)[1] <- "counts"
}

```

## After colData curation, remove samples with less than 2 M reads per sample

```{r}
parkinsonsMetaAnalysisData_pruned.tse <- parkinsonsMetaAnalysisData_pruned.tse[,parkinsonsMetaAnalysisData_pruned.tse@colData$reads_processed >= 2e6]
parkinsonsMetaAnalysisData_pruned.tse <- parkinsonsMetaAnalysisData_pruned.tse[rowSums(assay(parkinsonsMetaAnalysisData_pruned.tse)) > 0,]

sample_filtering_steps[["Min. 2M reads processed"]] <- parkinsonsMetaAnalysisData_pruned.tse@colData$uuid
```

## Write exclusions as json (list of uuids)

```{r}
write_json(sample_filtering_steps, path = sprintf("%s/00-pMD_%s_sample_filtering_steps.json", output_data.path, pipeline))
```

## Write TSE file

```{r, message=TRUE}

fullOutName <- sprintf("%s/pMD_%s.tse", output_data.path, pipeline)

dir.create(fullOutName, showWarnings = FALSE)

if(params$overwrite_TSE_out == "yes"){
  unlink(fullOutName, recursive = TRUE)
}

write_TSE_to_dir(parkinsonsMetaAnalysisData_pruned.tse, out.dir = fullOutName)

message(sprintf("Data saved as: %s", fullOutName))
```

## Check if just-written TSE is identical to the one in memory

It is basically identical, but `data.table::fwrite` has a precision of `1e-15`.
(see (#3862)[https://github.com/Rdatatable/data.table/issues/3862]).
In practice this precision is enough for bioinformatic analyses.

```{r}
re_read_tse <- read_TSE_from_dir(fullOutName)

all.equal(parkinsonsMetaAnalysisData_pruned.tse, re_read_tse, tolerance = 1e-16)
```

# Session Info

```{r}
sessionInfo()
```
