---
title: "Prepare pre-downloaded non-public data"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
 biobakery_tool: "MetaPhlAn"
 overwrite_TSE_out: "yes"
---

```{r setup}
if(!(params$biobakery_tool %in% c("MetaPhlAn", "HUMAnN"))){
  stop("Biobakery tool requested not existent, choose between 'MetaPhlAn' and 'HUMAnN'")
}

library(knitr) # CRAN
library(mia) # Bioconductor
library(tidyverse) # CRAN
library(magrittr) # CRAN
library(DT) # CRAN
library(jsonlite) # CRAN
library(janitor) # CRAN

# load or install install github libraries
library(parkinsonsMetagenomicData) # remotes::install_github("ASAP-MAC/parkinsonsMetagenomicData")
library(biobakeryUtils) # remotes::install_github("g-antonello/biobakeryUtils")
library(ChocoPhlAnTaxonomies) # remotes::install_github("g-antonello/ChocoPhlAnTaxonomies")


opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
sample_filtering_steps <- list()
output_data.path <- "~/parkinsonMetaAnalysis_data/02-private_local_datasets/"
```

```{r dataset name mapping vectors}
mapping_names.df <- data.frame(
  data_type = c("relative_abundance", "pathabundance_unstratified"),
  biobakery_tool = c("MetaPhlAn", "HUMAnN"),
  output_name = c("relative_abundance", "pathways")  
)

tool <- params$biobakery_tool
output_name <- mapping_names.df$output_name[mapping_names.df$biobakery_tool == tool]
data_type <- mapping_names.df$data_type[mapping_names.df$biobakery_tool == tool]
```

# Blue Poo dataset

The biobakery data of the Blue Poo dataset are public and available through
`parkinsonsMetagenomicData`. However, the metadata is private and should not 
leave the Unitn server.

## load pMD dataset `AsnicarF_2021`

```{r download AsnicarF_2021 from parkinsonsMetagenomicData}
data("sampleMetadata", "parkinsonsMetagenomicData")

AsnicarF_2021.tse <- sampleMetadata[sampleMetadata$study_name == "AsnicarF_2021",] %>% 
  returnSamples(data_type = data_type)

sample_filtering_steps[["Unfiltered"]] <- AsnicarF_2021.tse@colData$uuid
```

```{r prepare private bluePoo metadata}
#---------- 2 setup variables
bluePoo.path <- file.path(output_data.path, "bluePoo", "0-private")
#get selected variables
selected_variables <- c("participant_id", "bowel_movement_frequency_changed_last_3_months", "bowel_movements_last_7_days", 
"bristol_stool_score_average_last_3_months", "transit time (days)")
#------------------------

#------------- Prepare metadata 
# load metadata and mapping file
private_metadata.df <- read_tsv(file.path(bluePoo.path, "metadata_tt.tsv"), col_select = selected_variables)
participant_id_mapping.df <- read_tsv(file.path(bluePoo.path, "P1__mapp_sample_names.tsv")) %>% 
  select(participant_id = username, sample_accession)

# merge metadata and mapping file
bluePoo_metadata.df <- merge(participant_id_mapping.df, private_metadata.df, by = "participant_id")
rownames(bluePoo_metadata.df) <- bluePoo_metadata.df$sample_accession

# create Asnicar 2021-like metadata categorical variables
bluePoo_metadata.df <- mutate(bluePoo_metadata.df,
                              `transit time (hours)` = `transit time (days)` * 24,
    transitTimeCut = cut(`transit time (hours)`, breaks = c(0, 14, 38, 58,max(`transit time (hours)` + 1e-6, na.rm = TRUE)), right = FALSE),
    transitTimeCutClasses = cut(`transit time (hours)`, breaks = c(0, 14, 38, 58,max(`transit time (hours)` + 1e-6, na.rm = TRUE)), right = FALSE, labels = paste0("C", 1:4)),
    transitTimeCutClassesDefinition =  case_when(
     transitTimeCutClasses == "C1" ~ "Fast",
     transitTimeCutClasses %in% c("C2","C3") ~ "Normal",
     transitTimeCutClasses == "C4" ~ "Slow"
     ) %>% factor(c("Normal", "Fast", "Slow"))) %>% 
  # clean all column names to remove spaces
  clean_names()
#-----------------------------

#----- filter tse to have only available individuals and merge metadata
AsnicarF_2021.tse <- AsnicarF_2021.tse[, match(rownames(bluePoo_metadata.df), AsnicarF_2021.tse@colData$AsnicarF_2021_uncurated_sample_id)]
if(!identical(AsnicarF_2021.tse@colData$AsnicarF_2021_uncurated_sample_id, rownames(bluePoo_metadata.df))){
  stop("Subset and reordering of bluePoo metadata did not work")
}

new_coldata <- colData(AsnicarF_2021.tse) %>% 
  as.data.frame() %>% 
  rename(tmp_merge = AsnicarF_2021_uncurated_sample_id) %>% 
  left_join(bluePoo_metadata.df %>% rename(tmp_merge = sample_accession), by = "tmp_merge") %>% 
  select(-tmp_merge, -participant_id)
rownames(new_coldata) <- new_coldata$uuid

if(!identical(rownames(new_coldata), rownames(colData(AsnicarF_2021.tse)))){
  stop("something went wrong with row reordering of metadata to merge into AsnicarF_2021")
}

colData(AsnicarF_2021.tse) <- DataFrame(new_coldata)

AsnicarF_2021.tse@colData$target_condition <- "None"

# record samples lost in metadata replacement
sample_filtering_steps[["Private metadata available"]] <- AsnicarF_2021.tse@colData$uuid
```

## colData curation

```{r further curate AsnicarF_2021 colData}
# convert age to numeric
AsnicarF_2021.tse@colData$age <- as.numeric(AsnicarF_2021.tse@colData$age)

# Binary Parkinson variable (controls may carry other diseases)
AsnicarF_2021.tse@colData$parkinson_for_analysis <- factor(ifelse(grepl("Parkinson", AsnicarF_2021.tse@colData$disease), 1, 0))

# Recode missing age_group
AsnicarF_2021.tse@colData$age_group <- factor(ifelse(
  is.na(AsnicarF_2021.tse@colData$age_group),
  "Unknown",
  AsnicarF_2021.tse@colData$age_group
))

# Number of diseases (including PD)
AsnicarF_2021.tse@colData$number_diseases <- ifelse(
  AsnicarF_2021.tse@colData$disease == "Healthy",
  0,
  str_count(AsnicarF_2021.tse@colData$disease, ";") + 1
)

# Country conversion per study
country_per_study.chr <- c(
  BedarfJR_2017   = "Germany",
  BoktorJC_2023   = "USA",
  BoktorJC_2023.2 = "USA",
  DuruIC_2024     = "Finland",
  JoS_2022        = "South Korea",
  LeeEJ_2024      = "South Korea",
  MaoL_2021       = "China",
  NishiwakiH_2024 = "Japan",
  QianY_2020      = "China",
  WallenZD_2022   = "Deep South USA",
  ZhangM_2023     = "China",
  NGRC            = "USA",
  UAB             = "Deep South USA"
)

AsnicarF_2021.tse@colData$Country <- country_per_study.chr[AsnicarF_2021.tse@colData$study_name]
```

## biobakery tool-specific curation

```{r AsnicarF improve TSE formatting specific to biobakery tool}
AsnicarF_2021.tse <- pMD_enhance(AsnicarF_2021.tse, data_type = data_type)

if(tool == "MetaPhlAn"){
  # reads_processed as numeric values  
  AsnicarF_2021.tse@colData$reads_processed <- as.numeric(gsub("#([0-9]+).*", 
    "\\1", AsnicarF_2021.tse@colData$reads_processed))
  
  # reorder and rename assays to counts; relative_abundance; pathways
  assays(AsnicarF_2021.tse) <- assays(AsnicarF_2021.tse)[c(3, 1, 2)]
}

if(tool == "HUMAnN"){
  # assume some reads processed from the abundances of the pathways
  AsnicarF_2021.tse@colData$reads_processed <- colSums(assay(AsnicarF_2021.tse))
  
  # rename first assay
  assayNames(AsnicarF_2021.tse)[1] <- "counts"
}

```

## Exclude low-reads samples

```{r AsnicarF exclude samples with low sample counts}
AsnicarF_2021.tse <- AsnicarF_2021.tse[, AsnicarF_2021.tse@colData$reads_processed >= 2e6]
sample_filtering_steps[["Min. 2M reads processed"]] <- AsnicarF_2021.tse@colData$uuid
```

## Exclude all-0 samples

```{r AsnicarF exclude samples without abundance data}
AsnicarF_2021.tse <- AsnicarF_2021.tse[, colSums(assay(AsnicarF_2021.tse)) > 0]
sample_filtering_steps[["Sample abundances not entirely zero"]] <- AsnicarF_2021.tse@colData$uuid
```

## Sample abundances not entirely UNCLASSIFIED|UNMAPPED|UNINTEGRATED

```{r AsnicarF exclude samples with only UNCLASSIFIED or UNMAPPED}
all_abunds_UN <- sapply(assays(AsnicarF_2021.tse), function(x){
  x2 <- colSums(x[!grepl("UN", rownames(x)),, drop = FALSE]) == 0
  return(x2)
})

samples_to_remove <- apply(all_abunds_UN, 1, any)

# remove samples
AsnicarF_2021.tse <- AsnicarF_2021.tse[, !samples_to_remove]

sample_filtering_steps[["Sample abundances not entirely UNCLASSIFIED|UNMAPPED|UNINTEGRATED"]] <- AsnicarF_2021.tse@colData$uuid
```

## Write exclusions as json (list of uuids)

```{r AsnicarF write samples retained in each exclusion step}
write_json(sample_filtering_steps, path = sprintf("%s/../00-AsnicarF_2021_%s_sample_filtering_steps.json", bluePoo.path, tool))
```

## Write BluePoo (AsnicarF_2021)

```{r write TSE of bluePoo}
AsnicarF_2021.tse <- AsnicarF_2021.tse[rowSums(assay(AsnicarF_2021.tse)) > 0, ]

fullOutName <- file.path(sprintf("%s/../AsnicarF_2021_%s.tse", bluePoo.path, tool))

if(params$overwrite_TSE_out == "yes"){
  unlink(fullOutName, recursive = TRUE)
}

write_TSE_to_dir(AsnicarF_2021.tse, out.dir = fullOutName)
```

# Payami not-yet-public data (NGRC and UAB datasets)

These data were run with the same tool but were not included pMD because 
the authors did not publish reads online yet. Having a collaboration, we had 
official access to the data.

```{r inizialize a new stepwise sample filtering list}
sample_filtering_steps <- list()
```

## Metadata curation

```{r curate Payami metadata}
PayamiData.path <- file.path(output_data.path, "Payami", "0-private")

colData.df <- read_tsv(file.path(PayamiData.path, "colData.df.tsv")) %>% 
  as.data.frame() %>% 
  # curate a few columns so that they merge correctly with pMD
  mutate(
    study_name = Dataset,
    age = Age_at_collection,
    age_group = factor(ifelse(age >= 65, "Elderly", "Adult"), levels = c("Adult", "Elderly", "Unknown")),
    control = ifelse(Case_status == "PD", "Case", "Study Control"),
    sex = ifelse(Sex == "M", "Male", "Female"),
    target_condition = "Parkinson Disease"
    )

# create the "disease" column using columns 40 to 47
disease.subs <- colData.df[,c(13,40:47)] %>% 
  mutate(Case_status = ifelse(Case_status == "PD", "Y", "N")) %>% 
  rename(`Parkinson Disease` = Case_status,
         `Irritable Bowel Syndrome` = IBS,
         `Inflammatory Bowel Disease` = IBS,
         `Small Intestinal Bacterial Overgrowth` = SIBO,
         `Ulcer` = Ulcer_past_3_months, 
         `Celiac Disease` = Celiac_disease,
         `Crohn's disease` = Crohns_disease
         )

for(colNm in names(disease.subs)){
  disease.subs[[colNm]] <- ifelse(disease.subs[[colNm]] == "Y" & !is.na(disease.subs[[colNm]]), colNm, "")
}

colData.df$disease <- apply(disease.subs, 1, function(x) paste(x[x!= ""], collapse = ";")) %>% ifelse(. == "", "Healthy", .)

# define parkinson variable
colData.df$parkinson_for_analysis <- factor(ifelse(grepl("Parkinson", colData.df$disease), 1, 0))

# assign rownames to the uuids
rownames(colData.df) <- colData.df$uuid
```

## prepare basic TSE in case of MetaPhlAn

```{r prepare MetaPhlAn TSE}
if(tool == "MetaPhlAn"){
  # load merged input table
  # NB: it is already at SGB level
  input_basic.df <- read_tsv(file.path(PayamiData.path, "relative_abundance.tsv"))
  # first thing fix the UNCLASSIFIED case
  input_basic.df$clade_name[input_basic.df$clade_name == "UNCLASSIFIED"] <- paste(paste0(names(biobakeryUtils:::all_taxonomy_levels), "UNCLASSIFIED"), collapse = "|")
  
  # correct bug manually: there are 2x t__EUK4932, this is unexpected. abundances
  # of this duplicated row should be summed because it is the same feature.
  # generalizing to all:
  input_basic_fixedDups.df <- input_basic.df %>%
    group_by(clade_name) %>%
    summarise(across(everything(), sum)) %>%
    ungroup() %>%
    as.data.frame()
  
  # make the basic assay data
  input_basic_assay.mtx <- input_basic_fixedDups.df %>%
    column_to_rownames("clade_name") %>%
    as.matrix()
  
  # make the rowData
  rowData.df <- input_basic_fixedDups.df[, 1, drop = FALSE] %>%
    separate_wider_delim(
      cols = all_of("clade_name"),
      delim = "|",
      names = unname(biobakeryUtils:::all_taxonomy_levels),
      too_few = "align_start"
    )
  rownames(rowData.df) <- rowData.df$SGB
  
  # also set the rownames to the assay, the order is not changed, no checks needed
  rownames(input_basic_assay.mtx) <- rownames(rowData.df)
  
  # match rownames and colnames so all are there and reorder
  intersect_samples <- intersect(rownames(colData.df), colnames(input_basic_assay.mtx))
  colData.df <- colData.df[intersect_samples, ]
  input_basic_assay.mtx <- input_basic_assay.mtx[, intersect_samples]
  
  # create real relative abundance
  relative_abundance.mtx <- apply(input_basic_assay.mtx, 2, function(x)
    x / sum(x))
  
  # Create counts matrix
  assay_counts.mtx <- sweep(relative_abundance.mtx,
                            2,
                            colData.df$reads_processed,
                            "*")
  
  NGRC_UAB.tse <- TreeSummarizedExperiment(
    assays = list(
      "counts" = assay_counts.mtx,
      "relative_abundance" = relative_abundance.mtx,
      "percent" = input_basic_assay.mtx
    ),
    colData = DataFrame(colData.df),
    rowData = DataFrame(rowData.df)
  )
}
```

## prepare basic TSE in case of MetaPhlAn

```{r prepare HUMAnN TSE}
if(tool == "HUMAnN"){
  input_basic.df <- read_tsv(file.path(PayamiData.path, "pathabundance_unstratified.tsv"))
  
  # make assay-like file
  input_basic_assay.df <- input_basic.df %>%
    column_to_rownames("pathway")
  
  # extract first column that is going to become the raw rowData.df
  rowData.df <- input_basic.df[, 1, drop = FALSE]
  rownames(rowData.df) <- rowData.df[[1]]
  
  # match rownames and colnames so all are there and reorder
  intersect_samples <- intersect(rownames(colData.df), colnames(input_basic_assay.df))
  colData.df <- colData.df[intersect_samples, ]
  input_basic_assay.df <- input_basic_assay.df[, intersect_samples]
  
  NGRC_UAB.tse <- TreeSummarizedExperiment(
    assays = list("abundance" = as.matrix(input_basic_assay.df)),
    colData = DataFrame(colData.df),
    rowData = DataFrame(rowData.df)
  ) %>% pMD_enhance_HUMAnN_pwy()
  
  assayNames(NGRC_UAB.tse)[1] <- "counts"
  
  # replace reads_processed
  NGRC_UAB.tse@colData$reads_processed <- colSums(assay(NGRC_UAB.tse))
}
```

## Exclude low-reads samples

```{r Payami exclude samples with low sample counts}
NGRC_UAB.tse <- NGRC_UAB.tse[, NGRC_UAB.tse@colData$reads_processed >= 2e6]
sample_filtering_steps[["Min. 2M reads processed"]] <- NGRC_UAB.tse@colData$uuid
```

## Exclude all-0 samples

```{r Payami exclude samples without abundance data}
NGRC_UAB.tse <- NGRC_UAB.tse[, colSums(assay(NGRC_UAB.tse)) > 0]
sample_filtering_steps[["Sample abundances not entirely zero"]] <- NGRC_UAB.tse@colData$uuid
```

## Sample abundances not entirely UNCLASSIFIED|UNMAPPED|UNINTEGRATED

```{r Payami exclude samples with only UNCLASSIFIED or UNMAPPED}
all_abunds_UN <- sapply(assays(NGRC_UAB.tse), function(x){
  x2 <- colSums(x[!grepl("UN", rownames(x)),, drop = FALSE]) == 0
  return(x2)
})

samples_to_remove <- apply(all_abunds_UN, 1, any)

# remove samples
NGRC_UAB.tse <- NGRC_UAB.tse[, !samples_to_remove]

sample_filtering_steps[["Sample abundances not entirely UNCLASSIFIED|UNMAPPED|UNINTEGRATED"]] <- NGRC_UAB.tse@colData$uuid
```

## Write exclusions as json (list of uuids)

```{r Payami write samples retained in each exclusion step}
write_json(sample_filtering_steps, path = sprintf("%s/../00-NGRC_UAB_%s_sample_filtering_steps.json", PayamiData.path, tool))
```

## Write PayamiH TSE

```{r write Payami TSE data}
NGRC_UAB.tse <- NGRC_UAB.tse[rowSums(assay(NGRC_UAB.tse)) > 0,]

fullOutName <- sprintf("%s/../NGRC_UAB_%s.tse", PayamiData.path, tool)
  
if(params$overwrite_TSE_out == "yes"){
  unlink(fullOutName, recursive = TRUE)
}

write_TSE_to_dir(NGRC_UAB.tse, out.dir = fullOutName)
```

# Session Info

```{r session info}
sessionInfo()
```
