---
title: "Prepare pre-downloaded non-public data"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
 biobakery_tool: "MetaPhlAn"
 overwrite_TSE_out: "yes"
---

```{r setup}
if(!(params$biobakery_tool %in% c("MetaPhlAn", "HUMAnN"))){
  stop("Biobakery tool requested not existent, choose between 'MetaPhlAn' and 'HUMAnN'")
}

library(knitr) # CRAN
library(mia) # Bioconductor
library(tidyverse) # CRAN
library(magrittr) # CRAN
library(DT) # CRAN
library(jsonlite) # CRAN
library(janitor) # CRAN

# load or install install github libraries
library(parkinsonsMetagenomicData) # remotes::install_github("ASAP-MAC/parkinsonsMetagenomicData")
library(biobakeryUtils) # remotes::install_github("g-antonello/biobakeryUtils")
library(ChocoPhlAnTaxonomies) # remotes::install_github("g-antonello/ChocoPhlAnTaxonomies")


opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

source("../utils/pMD_data_prep_helpers.R")

sample_filtering_steps <- list()
output_data.path <- "~/parkinsonMetaAnalysis_data/02-private_local_datasets/"
```

```{r dataset name mapping vectors}
tool <- params$biobakery_tool
output_name <- biobakery_tool_map$output_name[biobakery_tool_map$biobakery_tool == tool]
data_type <- biobakery_tool_map$data_type[biobakery_tool_map$biobakery_tool == tool]
```

# Blue Poo dataset

The biobakery data of the Blue Poo dataset are public and available through
`parkinsonsMetagenomicData`. However, the metadata is private and should not 
leave the Unitn server.

## load pMD dataset `AsnicarF_2021`

```{r download AsnicarF_2021 from parkinsonsMetagenomicData}
data("sampleMetadata", package = "parkinsonsMetagenomicData")

AsnicarF_2021.tse <- sampleMetadata[sampleMetadata$study_name == "AsnicarF_2021",] %>% 
  returnSamples(data_type = data_type)

sample_filtering_steps[["Unfiltered"]] <- AsnicarF_2021.tse@colData$uuid
```

```{r prepare private bluePoo metadata}
#---------- 2 setup variables
bluePoo.path <- file.path(output_data.path, "bluePoo", "0-private")
#get selected variables
selected_variables <- c("participant_id", "bowel_movement_frequency_changed_last_3_months", "bowel_movements_last_7_days", 
"bristol_stool_score_average_last_3_months", "transit time (days)")
#------------------------

#------------- Prepare metadata 
# load metadata and mapping file
private_metadata.df <- read_tsv(file.path(bluePoo.path, "metadata_tt.tsv"), col_select = selected_variables)
participant_id_mapping.df <- read_tsv(file.path(bluePoo.path, "P1__mapp_sample_names.tsv")) %>% 
  select(participant_id = username, sample_accession)

# merge metadata and mapping file
bluePoo_metadata.df <- merge(participant_id_mapping.df, private_metadata.df, by = "participant_id")
rownames(bluePoo_metadata.df) <- bluePoo_metadata.df$sample_accession

# create Asnicar 2021-like metadata categorical variables
bluePoo_metadata.df <- mutate(bluePoo_metadata.df,
                              `transit time (hours)` = `transit time (days)` * 24,
    transitTimeCut = cut(`transit time (hours)`, breaks = c(0, 14, 38, 58,max(`transit time (hours)` + 1e-6, na.rm = TRUE)), right = FALSE),
    transitTimeCutClasses = cut(`transit time (hours)`, breaks = c(0, 14, 38, 58,max(`transit time (hours)` + 1e-6, na.rm = TRUE)), right = FALSE, labels = paste0("C", 1:4)),
    transitTimeCutClassesDefinition =  case_when(
     transitTimeCutClasses == "C1" ~ "Fast",
     transitTimeCutClasses %in% c("C2","C3") ~ "Normal",
     transitTimeCutClasses == "C4" ~ "Slow"
     ) %>% factor(c("Normal", "Fast", "Slow"))) %>% 
  # clean all column names to remove spaces
  clean_names()
#-----------------------------

#----- filter tse to have only available individuals and merge metadata
AsnicarF_2021.tse <- AsnicarF_2021.tse[, match(rownames(bluePoo_metadata.df), AsnicarF_2021.tse@colData$AsnicarF_2021_uncurated_sample_id)]
if(!identical(AsnicarF_2021.tse@colData$AsnicarF_2021_uncurated_sample_id, rownames(bluePoo_metadata.df))){
  stop("Subset and reordering of bluePoo metadata did not work")
}

new_coldata <- colData(AsnicarF_2021.tse) %>% 
  as.data.frame() %>% 
  rename(tmp_merge = AsnicarF_2021_uncurated_sample_id) %>% 
  left_join(bluePoo_metadata.df %>% rename(tmp_merge = sample_accession), by = "tmp_merge") %>% 
  select(-tmp_merge, -participant_id)
rownames(new_coldata) <- new_coldata$uuid

if(!identical(rownames(new_coldata), rownames(colData(AsnicarF_2021.tse)))){
  stop("something went wrong with row reordering of metadata to merge into AsnicarF_2021")
}

colData(AsnicarF_2021.tse) <- DataFrame(new_coldata)

# fix target condition in metadata for merging later
AsnicarF_2021.tse@colData$target_condition <- "None"
# Manually add SRA and BioProject 
AsnicarF_2021.tse@colData$BioProject <- "PRJEB39223"

# record samples lost in metadata replacement
sample_filtering_steps[["Private metadata available"]] <- AsnicarF_2021.tse@colData$uuid
```

## colData curation

```{r further curate AsnicarF_2021 colData}
# convert age to numeric
AsnicarF_2021.tse@colData$age <- as.numeric(AsnicarF_2021.tse@colData$age)

# Binary Parkinson variable (controls may carry other diseases)
AsnicarF_2021.tse@colData$parkinson_for_analysis <- factor("bluePoo")

# Recode missing age_group
AsnicarF_2021.tse@colData$age_group <- factor(ifelse(
  is.na(AsnicarF_2021.tse@colData$age_group),
  "Unknown",
  AsnicarF_2021.tse@colData$age_group
))

# Number of diseases (including PD)
AsnicarF_2021.tse@colData$number_diseases <- ifelse(
  AsnicarF_2021.tse@colData$disease == "Healthy",
  0,
  str_count(AsnicarF_2021.tse@colData$disease, ";") + 1
)

# Country conversion per study
AsnicarF_2021.tse@colData$Country <- country_per_study.chr[AsnicarF_2021.tse@colData$study_name]
```

## biobakery tool-specific curation

```{r AsnicarF improve TSE formatting specific to biobakery tool}
AsnicarF_2021.tse <- format_tool_specific_tse(AsnicarF_2021.tse, tool = tool, data_type = data_type)
```

## Exclude low-reads samples

```{r AsnicarF exclude samples with low sample counts}
AsnicarF_2021.tse <- AsnicarF_2021.tse[, AsnicarF_2021.tse@colData$reads_processed >= 2e6]
sample_filtering_steps[["Min. 2M reads processed"]] <- AsnicarF_2021.tse@colData$uuid
```

## Exclude all-0 samples

```{r AsnicarF exclude samples without abundance data}
AsnicarF_2021.tse <- AsnicarF_2021.tse[, colSums(assay(AsnicarF_2021.tse)) > 0]
sample_filtering_steps[["Sample abundances not entirely zero"]] <- AsnicarF_2021.tse@colData$uuid
```

## Sample abundances not entirely UNCLASSIFIED|UNMAPPED|UNINTEGRATED

```{r AsnicarF exclude samples with only UNCLASSIFIED or UNMAPPED}
all_abunds_UN <- sapply(assays(AsnicarF_2021.tse), function(x){
  x2 <- colSums(x[!grepl("UN", rownames(x)),, drop = FALSE]) == 0
  return(x2)
})

samples_to_remove <- apply(all_abunds_UN, 1, any)

# remove samples
AsnicarF_2021.tse <- AsnicarF_2021.tse[, !samples_to_remove]

sample_filtering_steps[["Sample abundances not entirely UNCLASSIFIED|UNMAPPED|UNINTEGRATED"]] <- AsnicarF_2021.tse@colData$uuid
```

## Write exclusions as json (list of uuids)

```{r AsnicarF write samples retained in each exclusion step}
write_json(sample_filtering_steps, path = sprintf("%s/../00-AsnicarF_2021_%s_sample_filtering_steps.json", bluePoo.path, tool))
```

## Write BluePoo (AsnicarF_2021)

```{r write TSE of bluePoo}
AsnicarF_2021.tse <- AsnicarF_2021.tse[rowSums(assay(AsnicarF_2021.tse)) > 0, ]

fullOutName <- file.path(sprintf("%s/../AsnicarF_2021_%s.tse", bluePoo.path, tool))

if(params$overwrite_TSE_out == "yes"){
  unlink(fullOutName, recursive = TRUE)
}

write_TSE_to_dir(AsnicarF_2021.tse, out.dir = fullOutName)
```

# Session Info

```{r session info}
sessionInfo()
```
