---
title: "Prepare pre-downloaded non-public data"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  pMD_data_type: "relative_abundance"
  overwrite_output: "yes"
  taxonomy_type: "GTDB"
---

```{r setup}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)

library(mia) # Bioconductor
library(tidyverse) # CRAN
library(magrittr) # CRAN
library(DT) # CRAN
library(janitor) # CRAN

# load or install install github libraries
library(parkinsonsMetagenomicData) # remotes::install_github("ASAP-MAC/parkinsonsMetagenomicData")
library(biobakeryUtils) # remotes::install_github("g-antonello/biobakeryUtils")
library(ChocoPhlAnTaxonomies) # remotes::install_github("g-antonello/ChocoPhlAnTaxonomies")
data("sampleMetadata")

private_data.path <- "~/parkinsonMetaAnalysis_data/01-private_local_datasets/"
```

# Blue Poo dataset

The biobakery data of the Blue Poo dataset are public and available through
`parkinsonsMetagenomicData`. However, the metadata is private and should not 
leave the Unitn server.

```{r}
AsnicarF_2021.tse <- sampleMetadata[sampleMetadata$study_name == "AsnicarF_2021",] %>% 
  returnSamples(data_type = params$pMD_data_type) %>% 
  pMD_enhance(data_type = params$pMD_data_type)
```

```{r}
#---------- 2 setup variables
bluePoo.path <- file.path(private_data.path, "bluePoo")
#get selected variables
selected_variables <- c("participant_id", "bowel_movement_frequency_changed_last_3_months", "bowel_movements_last_7_days", 
"bristol_stool_score_average_last_3_months", "transit time (days)")
#------------------------

#------------- Prepare metadata 
# load metadata and mapping file
private_metadata.df <- read_tsv(file.path(bluePoo.path, "metadata_tt.tsv"), col_select = selected_variables)
participant_id_mapping.df <- read_tsv(file.path(bluePoo.path, "P1__mapp_sample_names.tsv")) %>% 
  select(participant_id = username, sample_accession)

# merge metadata and mapping file
bluePoo_metadata.df <- merge(participant_id_mapping.df, private_metadata.df, by = "participant_id")
rownames(bluePoo_metadata.df) <- bluePoo_metadata.df$sample_accession

# create Asnicar 2021-like metadata categorical variables
bluePoo_metadata.df <- mutate(bluePoo_metadata.df,
                              `transit time (hours)` = `transit time (days)` * 24,
    transitTimeCut = cut(`transit time (hours)`, breaks = c(0, 14, 38, 58,max(`transit time (hours)` + 1e-6, na.rm = TRUE)), right = FALSE),
    transitTimeCutClasses = cut(`transit time (hours)`, breaks = c(0, 14, 38, 58,max(`transit time (hours)` + 1e-6, na.rm = TRUE)), right = FALSE, labels = paste0("C", 1:4)),
    transitTimeCutClassesDefinition =  case_when(
     transitTimeCutClasses == "C1" ~ "Fast",
     transitTimeCutClasses %in% c("C2","C3") ~ "Normal",
     transitTimeCutClasses == "C4" ~ "Slow"
     ) %>% factor(c("Normal", "Fast", "Slow"))) %>% 
  # clean all column names to remove spaces
  clean_names()
#-----------------------------

#----- filter tse to have only available individuals and merge metadata
AsnicarF_2021.tse <- AsnicarF_2021.tse[, match(rownames(bluePoo_metadata.df), AsnicarF_2021.tse@colData$AsnicarF_2021_uncurated_sample_id)]
if(!identical(AsnicarF_2021.tse@colData$AsnicarF_2021_uncurated_sample_id, rownames(bluePoo_metadata.df))){
  stop("Subset and reordering of bluePoo metadata did not work")
}
AsnicarF_2021.tse@colData <- cbind(colData(AsnicarF_2021.tse), select(bluePoo_metadata.df, -participant_id, sample_accession))

# correct metadata items and/or their classes
AsnicarF_2021.tse@colData$age <- as.numeric(AsnicarF_2021.tse@colData$age)
AsnicarF_2021.tse@colData$reads_processed <- as.numeric(AsnicarF_2021.tse@colData$reads_processed)
AsnicarF_2021.tse@colData$target_condition <- "None"
#-----------------------------
# final step: replace taxonomy with GTDB
if((params$pMD_data_type == "relative_abundance") & (params$taxonomy_type == "GTDB")){
AsnicarF_2021.tse <- tse_replace_mpa_with_GTDB_taxonomy(AsnicarF_2021.tse, chocophlan_datestamp = 202403)
}
```

## Write BluePoo (AsnicarF_2021)

```{r}
AsnicarF_2021.tse <- AsnicarF_2021.tse[rowSums(assay(AsnicarF_2021.tse))>0, ]
```

```{r}
fullOutName <- file.path(sprintf("~/parkinsonMetaAnalysis_data/02-intermediate-datasets/AsnicarF_2021_%s.tse", params$pMD_data_type))

if(params$overwrite_output == "yes"){
  unlink(fullOutName, recursive = TRUE)
}

write_TSE_to_dir(AsnicarF_2021.tse, out.dir = fullOutName)
```

# Payami not-yet-public data

These data were run with the same pipeline but were not included pMD because 
the authors did not publish reads online yet. Having a collaboration, we had 
official access to the data.

## NGRC + UAB Metadata curation

```{r}
PayamiData.path <- file.path(private_data.path, "MurchisonCF_2026", "input_from_cheaha")

colData.df <- read_tsv(file.path(PayamiData.path, "colData.df.tsv")) %>% 
  as.data.frame() %>% 
  # curate a few columns so that they merge correctly with pMD
  mutate(
    study_name = Dataset,
    age = Age_at_collection,
    age_group = ifelse(age >= 65, "Elderly", "Adult"),
    control = ifelse(Case_status == "PD", "Case", "Study Control"),
    sex = ifelse(Sex == "M", "Male", "Female"),
    target_condition = "Parkinson Disease"
    )

# create the disease column using columns 40 to 47
disease.subs <- colData.df[,c(13,40:47)] %>% 
  mutate(Case_status = ifelse(Case_status == "PD", "Y", "N")) %>% 
  rename(`Parkinson Disease` = Case_status,
         `Irritable Bowel Syndrome` = IBS,
         `Inflammatory Bowel Disease` = IBS,
         `Small Intestinal Bacterial Overgrowth` = SIBO,
         `Ulcer` = Ulcer_past_3_months, 
         `Celiac Disease` = Celiac_disease,
         `Crohn's disease` = Crohns_disease
         )

for(colNm in names(disease.subs)){
  disease.subs[[colNm]] <- ifelse(disease.subs[[colNm]] == "Y" & !is.na(disease.subs[[colNm]]), colNm, "")
}
# create variable "disease"
colData.df$disease <- apply(disease.subs, 1, function(x) paste(x[x!= ""], collapse = ";")) %>% ifelse(. == "", "Healthy", .)

# assign rownames to the uuids
rownames(colData.df) <- colData.df$uuid
```

## Prepare NGRC

```{r}
colData.df_sub <- colData.df %>% 
  filter(study_name =="NGRC")
```

### In case of HUMAnN `pathabundance_unstratified`

```{r eval=params$pMD_data_type == "pathabundance_unstratified"}
# load humann raw input
input_basic.df <- read_tsv(file.path(PayamiData.path, "pathabundance_unstratified.tsv"))

# make assay-like file
input_basic_assay.df <- input_basic.df %>% 
  column_to_rownames("pathway")

# extract first column that is going to become the raw rowData.df
rowData.df <- input_basic.df[,1, drop = FALSE]
rownames(rowData.df) <- rowData.df[[1]]
# match rownames and colnames so all are there and reorder
intersect_samples <- intersect(rownames(colData.df_sub), colnames(input_basic_assay.df))
colData.df_sub <- colData.df_sub[intersect_samples,]
input_basic_assay.df <- input_basic_assay.df[, intersect_samples]

NGRC.tse <- TreeSummarizedExperiment(
  assays = list("relative_abundance" = as.matrix(input_basic_assay.df)),
  colData = DataFrame(colData.df_sub),
  rowData = DataFrame(rowData.df)
) %>% pMD_enhance_HUMAnN_pwy()

```

### In case of MetaPhlAn `relative_abundance`

```{r, eval=params$pMD_data_type == "relative_abundance"}
# load merged input table
# NB: it is already at SGB level
input_basic.df <- read_tsv(file.path(PayamiData.path, "relative_abundance.tsv"))
# first thing fix the UNCLASSIFIED case
input_basic.df$clade_name[input_basic.df$clade_name == "UNCLASSIFIED"] <- paste(paste0(names(biobakeryUtils:::all_taxonomy_levels), "UNCLASSIFIED"), collapse = "|")

# correct bug manually: there are 2x t__EUK4932, this is unexpected. abundances 
# of this duplicated row should be summed because it is the same feature.
# generalizing to all:
input_basic_fixedDups.df <- input_basic.df %>%
  group_by(clade_name) %>%
  summarise(across(everything(), sum)) %>% 
  ungroup() %>% 
  as.data.frame()

# make the basic assay data
input_basic_assay.mtx <- input_basic_fixedDups.df %>% 
  column_to_rownames("clade_name") %>% 
  as.matrix()

# make the rowData 
rowData.df <- input_basic_fixedDups.df[, 1, drop = FALSE] %>%
  separate_wider_delim(
    cols = all_of("clade_name"),
    delim = "|",
    names = unname(biobakeryUtils:::all_taxonomy_levels),
    too_few = "align_start"
  )
rownames(rowData.df) <- rowData.df$SGB

# also set the rownames to the assay, the order is not changed, no checks needed
rownames(input_basic_assay.mtx) <- rownames(rowData.df)

# match rownames and colnames so all are there and reorder
intersect_samples <- intersect(rownames(colData.df_sub), colnames(input_basic_assay.mtx))
colData.df_sub <- colData.df_sub[intersect_samples,]
input_basic_assay.mtx <- input_basic_assay.mtx[, intersect_samples]

# create real relative abundance
relative_abundance.mtx <- apply(input_basic_assay.mtx, 2, function(x) x/sum(x))

# Create counts matrix
assay_counts.mtx <- sweep(relative_abundance.mtx, 2, colData.df_sub$reads_processed, "*")

NGRC.tse <- TreeSummarizedExperiment(
  assays = list("relative_abundance" = relative_abundance.mtx, "percent" = input_basic_assay.mtx, "counts" = assay_counts.mtx),
  colData = DataFrame(colData.df_sub),
  rowData = DataFrame(rowData.df)
)

# final step: replace taxonomy with GTDB
if((params$pMD_data_type == "relative_abundance") & (params$taxonomy_type == "GTDB")){
NGRC.tse <- tse_replace_mpa_with_GTDB_taxonomy(NGRC.tse, chocophlan_datestamp = 202403)
}
```

### Write NGRC

```{r}
fullOutName <- sprintf("~/parkinsonMetaAnalysis_data/02-intermediate-datasets/NGRC_%s.tse", params$pMD_data_type)
  
if(params$overwrite_output == "yes"){
  unlink(fullOutName, recursive = TRUE)
}

write_TSE_to_dir(NGRC.tse, out.dir = fullOutName)
```

## Prepare UAB

```{r}
colData.df_sub <- colData.df %>% 
  filter(study_name =="UAB")
```

### In case of `pathabundance_unstratified`

```{r eval=params$pMD_data_type == "pathabundance_unstratified"}
# load humann raw input
input_basic.df <- read_tsv(file.path(PayamiData.path, "pathabundance_unstratified.tsv"))

# make assay-like file
input_basic_assay.df <- input_basic.df %>% 
  column_to_rownames("pathway")

# extract first column that is going to become the raw rowData.df
rowData.df <- input_basic.df[,1, drop = FALSE]
rownames(rowData.df) <- rowData.df[[1]]
# match rownames and colnames so all are there and reorder
intersect_samples <- intersect(rownames(colData.df_sub), colnames(input_basic_assay.df))
colData.df_sub <- colData.df_sub[intersect_samples,]
input_basic_assay.df <- input_basic_assay.df[, intersect_samples]

UAB.tse <- TreeSummarizedExperiment(
  assays = list("relative_abundance" = as.matrix(input_basic_assay.df)),
  colData = DataFrame(colData.df_sub),
  rowData = DataFrame(rowData.df)
) %>% pMD_enhance_HUMAnN_pwy()
```

### In case of MetaPhlAn `relative_abundance`

```{r, eval=params$pMD_data_type == "relative_abundance"}
# load merged input table
# NB: it is already at SGB level
input_basic.df <- read_tsv(file.path(PayamiData.path, "relative_abundance.tsv"))
# first thing fix the UNCLASSIFIED case
input_basic.df$clade_name[input_basic.df$clade_name == "UNCLASSIFIED"] <- paste(paste0(names(biobakeryUtils:::all_taxonomy_levels), "UNCLASSIFIED"), collapse = "|")

# correct bug manually: there are 2x t__EUK4932, this is unexpected. abundances 
# of this duplicated row should be summed because it is the same feature.
# generalizing to all:
input_basic_fixedDups.df <- input_basic.df %>%
  group_by(clade_name) %>%
  summarise(across(everything(), sum)) %>% 
  ungroup() %>% 
  as.data.frame()

# make the basic assay data
input_basic_assay.mtx <- input_basic_fixedDups.df %>% 
  column_to_rownames("clade_name") %>% 
  as.matrix()

# make the rowData 
rowData.df <- input_basic_fixedDups.df[, 1, drop = FALSE] %>%
  separate_wider_delim(
    cols = all_of("clade_name"),
    delim = "|",
    names = unname(biobakeryUtils:::all_taxonomy_levels),
    too_few = "align_start"
  )
rownames(rowData.df) <- rowData.df$SGB

# also set the rownames to the assay, the order is not changed, no checks needed
rownames(input_basic_assay.mtx) <- rownames(rowData.df)

# match rownames and colnames so all are there and reorder
intersect_samples <- intersect(rownames(colData.df_sub), colnames(input_basic_assay.mtx))
colData.df_sub <- colData.df_sub[intersect_samples,]
input_basic_assay.mtx <- input_basic_assay.mtx[, intersect_samples]

# create real relative abundance
relative_abundance.mtx <- apply(input_basic_assay.mtx, 2, function(x) x/sum(x))

# Create counts matrix
assay_counts.mtx <- sweep(relative_abundance.mtx, 2, colData.df_sub$reads_processed, "*")

UAB.tse <- TreeSummarizedExperiment(
  assays = list("relative_abundance" = relative_abundance.mtx, "percent" = input_basic_assay.mtx, "counts" = assay_counts.mtx),
  colData = DataFrame(colData.df_sub),
  rowData = DataFrame(rowData.df)
)
# final step: replace taxonomy with GTDB
if((params$pMD_data_type == "relative_abundance") & (params$taxonomy_type == "GTDB")){
  UAB.tse <- tse_replace_mpa_with_GTDB_taxonomy(UAB.tse, chocophlan_datestamp = 202403)
}
```

### Write UAB

```{r}
UAB.tse <- UAB.tse[rowSums(assay(UAB.tse)) > 0,]
```

```{r}
fullOutName <- sprintf("~/parkinsonMetaAnalysis_data/02-intermediate-datasets/UAB_%s.tse", params$pMD_data_type)
  
if(params$overwrite_output == "yes"){
  unlink(fullOutName, recursive = TRUE)
}

write_TSE_to_dir(UAB.tse, out.dir = fullOutName)
```

# Session Info

```{r session info}
sessionInfo()
```
