---
title: "Prepare pre-downloaded non-public data"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_tool: "MetaPhlAn"
  overwrite_TSE_out: "yes"
  preval_thresh: 0.05
  detect_thresh: 1e-5
---

```{r setup}
if(!(params$biobakery_tool %in% c("MetaPhlAn", "HUMAnN"))){
  stop("Biobakery tool requested not existent, choose between 'MetaPhlAn' and 'HUMAnN'")
}

library(knitr)
library(mia)
library(biobakeryUtils)
library(ChocoPhlAnTaxonomies)
library(tidyverse)

opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
sample_filtering_steps <- list()
output_data.path <- "~/parkinsonMetaAnalysis_data/"
```

```{r dataset name mapping vectors}
mapping_names.df <- data.frame(
  data_type = c("relative_abundance", "pathabundance_unstratified"),
  biobakery_tool = c("MetaPhlAn", "HUMAnN"),
  output_name = c("relative_abundance", "pathways")  
)

tool <- params$biobakery_tool
output_name <- mapping_names.df$output_name[mapping_names.df$biobakery_tool == tool]
data_type <- mapping_names.df$data_type[mapping_names.df$biobakery_tool == tool]
```

```{r load and merge datasets}
# load datasets
PD_data.list <- list(
  pMD = read_TSE_from_dir(
    sprintf(sprintf("%s/01-pMD_datasets/pMD_%s.tse", output_data.path, tool))
  ),
  NGRC_UAB = read_TSE_from_dir(
    sprintf("%s/02-private_local_datasets/Payami/NGRC_UAB_%s.tse/", output_data.path, tool)
  ),
  AsnicarF_2021.tse = read_TSE_from_dir(sprintf("%s/02-private_local_datasets/bluePoo/AsnicarF_2021_%s.tse", output_data.path, tool))
) %>% 
  # workaround on age being character, not sure why this still happens
  lapply(function(x){
   x$age <- as.numeric(x$age)
    return(x)
  }
    )

# Merge all datasets counts
tmp_counts.tse <- suppressMessages(reduce(
  PD_data.list,
  mergeSEs,
  assay.type =  "counts",
  missing.values = 0,
  collapse.rows  = TRUE
))

# merge all dataset relative_abundances
tmp_relative_abundance.tse <- suppressMessages(reduce(
  PD_data.list,
  mergeSEs,
  assay.type =  "relative_abundance",
  missing.values = 0,
  collapse.rows  = TRUE
))

# transfer relative_abundance into main merged tse
all_data_merged.tse <- tmp_counts.tse
assay(all_data_merged.tse, "relative_abundance") <- assay(tmp_relative_abundance.tse)

# MetaPhlAn specific steps
if(params$biobakery_tool == "MetaPhlAn"){
  # Fix column name casing that mia changes on the SGB column in case of MetaPhlAn
  colnames(rowData(all_data_merged.tse))[8] <- toupper(colnames(rowData(all_data_merged.tse))[8])
  # replace taxonomy with GTDB taxonomy
  all_data_merged.tse <- ChocoPhlAnTaxonomies::tse_replace_mpa_with_GTDB_taxonomy(all_data_merged.tse, 202403)
}
```

# Filter by prevalence

Before filter

```{r print dataset before any filter is applied}
all_data_merged.tse
```

After filter

```{r print dataset after prevalence and abundance filters are applied}
Input.tse <- subsetByPrevalent(
  all_data_merged.tse,
  assay.type = "relative_abundance",
  prevalence = params$preval_thresh,
  detection  = params$detect_thresh
)

Input.tse
```

## Generate `constipation_for_analysis` variable

This had to be done now because 3 out of 5 datasets to analyze constipation
were not available before now.

**AsnicarF_2021** is the gold standard, with a pre-set definition: `Normal`, 
`Fast`, `Slow`.

**Payami** private datasets have Constipation defined as yes/no answer to the 
question "have you been constipated in the 3 months before stool collection?".
Here, the values are converted as:
  - Slow: Constipation == "Y"
  - Normal: Constipation == "N"
  - Fast: NA (not defined by the variable)

**NishiwakiH_2024** also has a proxy for constipation, `stool_frequency`, 
measured as number of bowel movements per week (BMPW). According to 
[Walter et al. (2010)](https://doi.org/10.3109/00365520903551332), this variable
can be cut as follows:
  
  - Slow: $BMPW < 3$
  - Normal: $3\leqslant BMPW \leqslant 21$
  - Fast: $BMPW \geq 21$

**BoktorJC_2023** has a Bristol stool scale variable. It can be categorized into
stool with different moisture. Less moisture tends to correlate with slow transit
time, which increases the readsorption.The variable to curate constipation is 
`BoktorJC_2023_uncurated_bristol_stool_scale2`, generate in the 01-... script.

This is derived from two uncurated variables.

  - BoktorJC_2023_uncurated_bristol_stool_scale
  - BoktorJC_2023_uncurated_Bristol_stool_scale
  
The first one is more reliable, described by researchers. In those few cases 
where the second variable has values and the 1st doesn't, values from the 2nd 
are taken.

Bristol stool scores can be grouped as follows according to slow or constipation
according to[Orozco et al., 2024](https://doi.org/10.14309/ajg.0000000000003218):

  - Slow (constipation): $Bristol \leqslant 2$
  - Normal: $3 \leqslant Bristol \leqslant 5$
  - Fast (diarrhea): $6 \leqslant Bristol \leqslant 7$

Boktor also has a `constipation_problems` coming from the UPDRS-II questionnaire.
It ranges from 0 (normal) to 4 (severe).

An overarching constipation variable is then derived from the two above, in which
if either reaches constipation levels, the sample is defined constipated.

Many variables with different scales and meanings are harmonized together. To 
be cautious on the definition resulting from this grouping, the Asnicar/bluePoo
definitions are applied.


```{r generate constipation variable}
# fix empty slots of WallenZD constipation variable
Input.tse$WallenZD_2022_uncurated_constipation[Input.tse$WallenZD_2022_uncurated_constipation == ""] <- NA 

# merge WallenZD uncurated constipation into the main Constipation variable
# coming from the Payami datasets (Wallen, NGRC, UAB)
Input.tse$Constipation <- case_when(is.na(Input.tse$Constipation) ~ Input.tse$WallenZD_2022_uncurated_constipation, TRUE ~ Input.tse$Constipation)

# create one intermediate constipation_payami variable
Input.tse$constipation_payami <- case_when(
  Input.tse$Constipation == "Y" ~ "Slow", 
  Input.tse$Constipation == "N" ~ "Normal", 
  TRUE ~ NA
  )

Input.tse$constipation_for_analysis <- case_when(
  is.na(Input.tse$Constipation) &
    is.na(Input.tse$transit_time_cut_classes_definition) ~ NA,
  Input.tse$Constipation == "Y" ~ "Slow",
  Input.tse$transit_time_cut_classes_definition == "Slow" ~ "Slow",
  Input.tse$Constipation == "N" ~ "Normal",
  Input.tse$transit_time_cut_classes_definition == "Normal" ~ "Normal",
  TRUE ~ "Fast"
)

# Create Boktor intermediate constipation into constipation_for_analysis
## 1 - group bristol scores
Input.tse$boktor_constipation_bristol <- case_when(
  Input.tse$BoktorJC_2023_uncurated_bristol_stool_scale2 <= 2 ~ "Slow",
  Input.tse$BoktorJC_2023_uncurated_bristol_stool_scale2 > 2 ~ "Normal",
  Input.tse$BoktorJC_2023_uncurated_bristol_stool_scale2 > 5 ~ "Fast",
  TRUE ~ Input.tse$BoktorJC_2023_uncurated_bristol_stool_scale2
)

## 2 - make an OR statement with variable just created and 
Input.tse$boktor_constipation_updrs <- case_when(
  as.integer(Input.tse$BoktorJC_2023_uncurated_constipation_problems) <= 1 ~ "Normal",
  as.integer(Input.tse$BoktorJC_2023_uncurated_constipation_problems) > 1 ~ "Slow",
  TRUE ~ NA
)

# coalesce Boktor constipation variables together
Input.tse$boktor_constipation <- coalesce(Input.tse$constipation_for_analysis, Input.tse$boktor_constipation_updrs)

# merge boktor into constipation_for_analysis
Input.tse$constipation_for_analysis <- coalesce(Input.tse$boktor_constipation_bristol, Input.tse$boktor_constipation)

# merge Nishiwaki stool_frequency into the main constipation_for_analysis:
## 1 - Curate nishiwaki constipation from the stool frequency
Input.tse$constipation_nishiwaki <-  case_when(
  as.numeric(Input.tse$NishiwakiH_2024_uncurated_stool_frequency) < 3 ~ "Slow",
  as.numeric(Input.tse$NishiwakiH_2024_uncurated_stool_frequency) >= 3 ~ "Normal",
  as.numeric(Input.tse$NishiwakiH_2024_uncurated_stool_frequency) >= 21 ~ "Fast",
  TRUE ~ NA)

# make the variable a factor with Normal as baseline
Input.tse$constipation_for_analysis <- coalesce(Input.tse$constipation_for_analysis, Input.tse$constipation_nishiwaki)

Input.tse$constipation_for_analysis <- factor(Input.tse$constipation_for_analysis, levels = c("Normal", "Fast", "Slow"))

kable(table(Input.tse$constipation_for_analysis, Input.tse$parkinson_for_analysis))
```

## Write the complete dataset 

```{r write complete dataset}
outputName <- sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_complete-%s.tse", tool)

if(params$overwrite_TSE_out == "yes"){
  unlink(outputName, recursive = TRUE)
}

write_TSE_to_dir(Input.tse, out.dir = outputName)
```

## Write a dataset to be used for the `Parkinson` meta-analysis

```{r filter dataset for parkinson meta-analysis}
# Parkinson as target condition
Input_parkinson.tse <- Input.tse[, Input.tse$target_condition == "Parkinson Disease"]
Input_parkinson.tse <- Input_parkinson.tse[rowSums(assay(Input_parkinson.tse)) > 0, ]

# Each dataset should have at least 30 samples in cases and 30 in controls
ok_datasets <- Input_parkinson.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  filter(
    sum(parkinson_for_analysis == 0) > 10,
    sum(parkinson_for_analysis == 1) > 10
  )

## filter TSE to restrict to 'ok' datasets 
Input_parkinson_usableDatasets.tse <- Input_parkinson.tse[,Input_parkinson.tse$study_name %in% unique(ok_datasets$study_name)]

## filter all-0 rows again
Input_parkinson_usableDatasets.tse <- Input_parkinson_usableDatasets.tse[rowSums(assay(Input_parkinson_usableDatasets.tse)) > 0,]
```


```{r write dataset for parkinson meta-analysis}
# save tse
outputName <- sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/02-dataset_PD-%s.tse", tool)

if(params$overwrite_TSE_out == "yes"){
  unlink(outputName, recursive = TRUE)
}
write_TSE_to_dir(Input_parkinson_usableDatasets.tse, out.dir = outputName)

Input_parkinson_usableDatasets.tse
```

## Write a dataset to be used for the `Healthy Constipated` meta-analysis

```{r filter dataset for constipation meta-analysis}
# filter for disease == Healthy. From 3384 to 1709
Input_constipation_healthy.tse <- Input.tse[, Input.tse$disease == "Healthy"]

# filter for available constipation information. From 1709 to 1242
Input_constipation_healthy.tse <- Input_constipation_healthy.tse[,!is.na(Input_constipation_healthy.tse$constipation_for_analysis)]

kable(table(Input_constipation_healthy.tse$study_name, Input_constipation_healthy.tse$constipation_for_analysis, useNA = "always"))
```

```{r write dataset for constipation meta-analysis}
outputName <- sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/03-dataset_constipation-%s.tse", tool)

if(params$overwrite_TSE_out == "yes"){
  unlink(outputName, recursive = TRUE)
}

write_TSE_to_dir(Input_constipation_healthy.tse, outputName)
```

# Session Info

```{r session info}
sessionInfo()
```
