---
title: "Prepare pre-downloaded non-public data"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_pipeline: "MetaPhlAn"
  overwrite_TSE_out: "yes"
  preval_thresh: 0.05
  detect_thresh: 1e-5
---

```{r setup}
if(!(params$biobakery_pipeline %in% c("MetaPhlAn", "HUMAnN"))){
  stop("Biobakery pipeline requested not existent, choose between 'MetaPhlAn' and 'HUMAnN'")
}

library(knitr)
library(mia)
library(biobakeryUtils)
library(tidyverse)

opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
sample_filtering_steps <- list()
output_data.path <- "~/parkinsonMetaAnalysis_data/"
```

```{r dataset name mapping utilities}
mapping_names.df <- data.frame(
  data_type = c("relative_abundance", "pathabundance_unstratified"),
  biobakery_pipeline = c("MetaPhlAn", "HUMAnN"),
  output_name = c("relative_abundance", "pathways")  
)

pipeline <- params$biobakery_pipeline
output_name <- mapping_names.df$output_name[mapping_names.df$biobakery_pipeline == pipeline]
data_type <- mapping_names.df$data_type[mapping_names.df$biobakery_pipeline == pipeline]
```

```{r load and merge datasets}
# load datasets
PD_data.list <- list(
  pMD = read_TSE_from_dir(
    sprintf(sprintf("%s/01-pMD_datasets/pMD_%s.tse", output_data.path, pipeline))
  ),
  NGRC_UAB = read_TSE_from_dir(
    sprintf("%s/02-private_local_datasets/Payami/NGRC_UAB_%s.tse/", output_data.path, pipeline)
  ),
  AsnicarF_2021.tse = read_TSE_from_dir(sprintf("%s/02-private_local_datasets/bluePoo/AsnicarF_2021_%s.tse", output_data.path, pipeline))
) %>% 
  # workaround on age being character, not sure why this still happens
  lapply(function(x){
   x@colData$age <- as.numeric(x@colData$age)
    return(x)
  }
    )

# Merge all datasets counts
tmp_counts.tse <- suppressMessages(reduce(
  PD_data.list,
  mergeSEs,
  assay.type =  "counts",
  missing.values = 0,
  collapse.rows  = TRUE
))

# merge all dataset relative_abundances
tmp_relative_abundance.tse <- suppressMessages(reduce(
  PD_data.list,
  mergeSEs,
  assay.type =  "relative_abundance",
  missing.values = 0,
  collapse.rows  = TRUE
))

# transfer relative_abundance into main merged tse
all_data_merged.tse <- tmp_counts.tse
assay(all_data_merged.tse, "relative_abundance") <- assay(tmp_relative_abundance.tse)

# Fix column name casing that mia changes on the SGB column in case of MetaPhlAn
if(params$biobakery_pipeline == "MetaPhlAn"){
colnames(rowData(all_data_merged.tse))[8] <- toupper(colnames(rowData(all_data_merged.tse))[8])
}
```

# Filter by prevalence

Before filter

```{r}
all_data_merged.tse
```

After filter

```{r}
Input.tse <- subsetByPrevalent(
  all_data_merged.tse,
  assay.type = "relative_abundance",
  prevalence = params$preval_thresh,
  detection  = params$detect_thresh
)

Input.tse
```

## Generate `constipation_for_analysis` variable

This had to be done now because 3 out of 4 datasets datasets to analyze constipation
were not available before now.

First curate a comprehensive constipation variable, this is knowledge based
work. I will take Asnicar2021 as gold standard and assume of Murchison datasets
are N = "Normal", Y = "Slow"

```{r}
# fix empty slots of WallenZD constipation variable
Input.tse@colData$WallenZD_2022_uncurated_constipation[Input.tse@colData$WallenZD_2022_uncurated_constipation == ""] <- NA 

# merge WallenZD uncurated constipation into the main Constipation variable
Input.tse@colData$Constipation <- case_when(is.na(Input.tse@colData$Constipation) ~ Input.tse@colData$WallenZD_2022_uncurated_constipation, TRUE ~ Input.tse@colData$Constipation)

Input.tse@colData$constipation_for_analysis <- case_when(is.na(Input.tse@colData$Constipation) & is.na(Input.tse@colData$transit_time_cut_classes_definition) ~ NA,
                                                   Input.tse@colData$Constipation == "Y" ~ "Slow",
                                                   Input.tse@colData$transit_time_cut_classes_definition == "Slow" ~ "Slow",
                                                   Input.tse@colData$Constipation == "N" ~ "Normal",
                                                   Input.tse@colData$transit_time_cut_classes_definition == "Normal" ~ "Normal",
                                                   TRUE ~ "Fast"
                                                   )

outputName <- sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_complete-%s.tse", pipeline)

if(params$overwrite_TSE_out == "yes"){
  unlink(outputName, recursive = TRUE)
}

write_TSE_to_dir(Input.tse, out.dir = outputName)
```

# Code checked until here

## Write a dataset to be used for the `Parkinson` meta-analysis

```{r}
# Parkinson as target condition
Input_parkinson.tse <- Input.tse[, Input.tse@colData$target_condition == "Parkinson Disease"]
Input_parkinson.tse <- Input_parkinson.tse[rowSums(assay(Input_parkinson.tse)) > 0, ]

# Each dataset should have at least 30 samples in cases and 30 in controls
ok_datasets <- Input_parkinson.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  filter(
    sum(parkinson_for_analysis == 0) > 30,
    sum(parkinson_for_analysis == 1) > 30
  )

## filter TSE to restrict to 'ok' datasets 
Input_parkinson_usableDatasets.tse <- Input_parkinson.tse[,Input_parkinson.tse@colData$study_name %in% unique(ok_datasets$study_name)]

## filter all-0 rows again
Input_parkinson_usableDatasets.tse <- Input_parkinson_usableDatasets.tse[rowSums(assay(Input_parkinson_usableDatasets.tse)) > 0,]


# save tse
outputName <- sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_PD-%s.tse", pipeline)

if(params$overwrite_TSE_out == "yes"){
  unlink(outputName, recursive = TRUE)
}
write_TSE_to_dir(Input_parkinson_usableDatasets.tse, out.dir = outputName)

Input_parkinson_usableDatasets.tse
```

# Session Info

```{r}
sessionInfo()
```

