---
title: "Prepare pre-downloaded non-public data"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  overwrite_TSE_out: "yes"
---

```{r setup}
library(knitr)
library(mia)
library(biobakeryUtils)
library(ChocoPhlAnTaxonomies)
library(tidyverse)

opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)
sample_filtering_steps <- list()
output_data.path <- "~/parkinsonMetaAnalysis_data/"
```

# Load and merge MetaPhlAn relative_abundance data

```{r load and merge MetaPhlAn datasets}
# load datasets
PD_data.list <- list(
  pMD = read_TSE_from_dir(
    file.path(output_data.path, "01-pMD_datasets/", "pMD_MetaPhlAn.tse")
  ),
  AsnicarF_2021.tse = read_TSE_from_dir(
    file.path(output_data.path, "02-private_local_datasets", "bluePoo", "AsnicarF_2021_MetaPhlAn.tse")
    )) %>% 
  # workaround on age being character, not sure why this still happens
  lapply(function(x){
   x$age <- as.numeric(x$age)
    return(x)
  }
    )

# Merge all datasets counts
tmp_counts.tse <- suppressMessages(reduce(
  PD_data.list,
  mergeSEs,
  assay.type =  "counts",
  missing.values = 0,
  collapse.rows  = TRUE
))

# merge all dataset relative_abundances
tmp_relative_abundance.tse <- suppressMessages(reduce(
  PD_data.list,
  mergeSEs,
  assay.type =  "relative_abundance",
  missing.values = 0,
  collapse.rows  = TRUE
))

# transfer relative_abundance into main merged tse
MetaPhlAn_data_merged.tse <- tmp_counts.tse
assay(MetaPhlAn_data_merged.tse, "relative_abundance") <- assay(tmp_relative_abundance.tse)

# Fix column name casing that mia changes on the SGB column in case of MetaPhlAn
colnames(rowData(MetaPhlAn_data_merged.tse))[8] <- toupper(colnames(rowData(MetaPhlAn_data_merged.tse))[8])

# replace taxonomy with GTDB taxonomy
MetaPhlAn_data_merged.tse <- ChocoPhlAnTaxonomies::tse_replace_mpa_with_GTDB_taxonomy(MetaPhlAn_data_merged.tse, 202403)
```

## Filter MetaPhlAn data to have at least 2M reads per sample

```{r}
MetaPhlAn_data_merged.tse <- MetaPhlAn_data_merged.tse[, MetaPhlAn_data_merged.tse$reads_processed >= 2e6]
```

## Generate `constipation_for_analysis` variable

This had to be done now because 3 out of 5 datasets to analyze constipation
were not available before now.

**AsnicarF_2021** is the gold standard, with a pre-set definition: `Normal`, 
`Fast`, `Slow`.

**NishiwakiH_2024** also has a proxy for constipation, `stool_frequency`, 
measured as number of bowel movements per week (BMPW). According to 
[Walter et al. (2010)](https://doi.org/10.3109/00365520903551332), this variable
can be cut as follows:
  
  - Slow: $BMPW < 3$
  - Normal: $3\leqslant BMPW \leqslant 21$
  - Fast: $BMPW \geq 21$

**BoktorJC_2023** has a Bristol stool scale variable. It can be categorized into
stool with different moisture. Less moisture tends to correlate with slow transit
time, which increases the readsorption.The variable to curate constipation is 
`BoktorJC_2023_uncurated_bristol_stool_scale2`, generate in the 01-... script.

This is derived from two uncurated variables.

  - BoktorJC_2023_uncurated_bristol_stool_scale
  - BoktorJC_2023_uncurated_Bristol_stool_scale
  
The first one is more reliable, described by researchers. In those few cases 
where the second variable has values and the 1st doesn't, values from the 2nd 
are taken.

Bristol stool scores can be grouped as follows according to slow or constipation
according to[Orozco et al., 2024](https://doi.org/10.14309/ajg.0000000000003218):

  - Slow (constipation): $Bristol \leqslant 2$
  - Normal: $3 \leqslant Bristol \leqslant 5$
  - Fast (diarrhea): $6 \leqslant Bristol \leqslant 7$

Boktor also has a `constipation_problems` coming from the UPDRS-II questionnaire.
It ranges from 0 (normal) to 4 (severe).

An overarching constipation variable is then derived from the two above, in which
if either reaches constipation levels, the sample is defined constipated.

Many variables with different scales and meanings are harmonized together. To 
be cautious on the definition resulting from this grouping, the Asnicar/bluePoo
definitions are applied.

```{r generate constipation variable}
# fix empty slots of WallenZD constipation variable
MetaPhlAn_data_merged.tse$WallenZD_2022_uncurated_constipation[MetaPhlAn_data_merged.tse$WallenZD_2022_uncurated_constipation == ""] <- NA 
MetaPhlAn_data_merged.tse$Constipation <- MetaPhlAn_data_merged.tse$WallenZD_2022_uncurated_constipation

MetaPhlAn_data_merged.tse$constipation_for_analysis <- case_when(
  is.na(MetaPhlAn_data_merged.tse$Constipation) &
    is.na(MetaPhlAn_data_merged.tse$transit_time_cut_classes_definition) ~ NA,
  MetaPhlAn_data_merged.tse$Constipation == "Y" ~ "Slow",
  MetaPhlAn_data_merged.tse$transit_time_cut_classes_definition == "Slow" ~ "Slow",
  MetaPhlAn_data_merged.tse$Constipation == "N" ~ "Normal",
  MetaPhlAn_data_merged.tse$transit_time_cut_classes_definition == "Normal" ~ "Normal",
  TRUE ~ "Fast"
)

# Create Boktor intermediate constipation into constipation_for_analysis
## 1 - group bristol scores
MetaPhlAn_data_merged.tse$constipation_boktor_bristol <- case_when(
  MetaPhlAn_data_merged.tse$BoktorJC_2023_uncurated_bristol_stool_scale2 <= 2 ~ "Slow",
  MetaPhlAn_data_merged.tse$BoktorJC_2023_uncurated_bristol_stool_scale2 > 2 ~ "Normal",
  MetaPhlAn_data_merged.tse$BoktorJC_2023_uncurated_bristol_stool_scale2 > 5 ~ "Fast",
  TRUE ~ MetaPhlAn_data_merged.tse$BoktorJC_2023_uncurated_bristol_stool_scale2
)

## 2 - make an OR statement with variable just created
MetaPhlAn_data_merged.tse$constipation_boktor_updrs <- case_when(
  as.integer(MetaPhlAn_data_merged.tse$BoktorJC_2023_uncurated_constipation_problems) <= 1 ~ "Normal",
  as.integer(MetaPhlAn_data_merged.tse$BoktorJC_2023_uncurated_constipation_problems) > 1 ~ "Slow",
  TRUE ~ NA
)

# coalesce Boktor constipation variables together
MetaPhlAn_data_merged.tse$constipation_boktor <- coalesce(MetaPhlAn_data_merged.tse$constipation_boktor_bristol, MetaPhlAn_data_merged.tse$constipation_boktor_updrs)

# merge boktor into constipation_for_analysis
MetaPhlAn_data_merged.tse$constipation_for_analysis <- coalesce(MetaPhlAn_data_merged.tse$constipation_for_analysis, MetaPhlAn_data_merged.tse$constipation_boktor)

# merge Nishiwaki stool_frequency into the main constipation_for_analysis:
## 1 - Curate nishiwaki constipation from the stool frequency
MetaPhlAn_data_merged.tse$constipation_nishiwaki <-  case_when(
  as.numeric(MetaPhlAn_data_merged.tse$NishiwakiH_2024_uncurated_stool_frequency) < 3 ~ "Slow",
  as.numeric(MetaPhlAn_data_merged.tse$NishiwakiH_2024_uncurated_stool_frequency) >= 3 ~ "Normal",
  as.numeric(MetaPhlAn_data_merged.tse$NishiwakiH_2024_uncurated_stool_frequency) >= 21 ~ "Fast",
  TRUE ~ NA)

# make the variable a factor with Normal as baseline
MetaPhlAn_data_merged.tse$constipation_for_analysis <- coalesce(MetaPhlAn_data_merged.tse$constipation_for_analysis, MetaPhlAn_data_merged.tse$constipation_nishiwaki)

MetaPhlAn_data_merged.tse$constipation_for_analysis <- factor(MetaPhlAn_data_merged.tse$constipation_for_analysis, levels = c("Normal", "Fast", "Slow"))

kable(table(MetaPhlAn_data_merged.tse$constipation_for_analysis, MetaPhlAn_data_merged.tse$parkinson_for_analysis))
```

# Load and merge Pathway data

```{r load and merge HUMAnN datasets}
# load datasets
PD_data.list <- list(
  pMD = read_TSE_from_dir(
    file.path(output_data.path, "01-pMD_datasets/", "pMD_HUMAnN.tse")
  ),
  AsnicarF_2021.tse = read_TSE_from_dir(
    file.path(output_data.path, "02-private_local_datasets", "bluePoo", "AsnicarF_2021_HUMAnN.tse")
    )) %>% 
  # workaround on age being character, not sure why this still happens
  lapply(function(x){
   x$age <- as.numeric(x$age)
    return(x)
  }
    )

# Merge all datasets counts
tmp_counts.tse <- suppressMessages(reduce(
  PD_data.list,
  mergeSEs,
  assay.type =  "counts",
  missing.values = 0,
  collapse.rows  = TRUE
))

# merge all dataset relative_abundances
tmp_relative_abundance.tse <- suppressMessages(reduce(
  PD_data.list,
  mergeSEs,
  assay.type =  "relative_abundance",
  missing.values = 0,
  collapse.rows  = TRUE
))

# transfer relative_abundance into main merged tse
HUMAnN_data_merged.tse <- tmp_counts.tse
assay(HUMAnN_data_merged.tse, "relative_abundance") <- assay(tmp_relative_abundance.tse)
```

## Replace metadata of Pathways with the newly generated metadata from MetaPhlAn

```{r}
metadata_to_replace <- colData(MetaPhlAn_data_merged.tse)[HUMAnN_data_merged.tse$uuid,]
colData(HUMAnN_data_merged.tse) <- metadata_to_replace
```

## Keep only samples available in both datasets

```{r}
sample_intersect <- intersect(MetaPhlAn_data_merged.tse$uuid, HUMAnN_data_merged.tse$uuid)
MetaPhlAn_data_merged.tse <- MetaPhlAn_data_merged.tse[, MetaPhlAn_data_merged.tse$uuid %in% sample_intersect]
HUMAnN_data_merged.tse <- HUMAnN_data_merged.tse[, HUMAnN_data_merged.tse$uuid %in% sample_intersect]
```

# Write the complete datasets

## MetaPhlAn

```{r write complete MetaPhlAn dataset}
outputName <- file.path("~", "parkinsonMetaAnalysis_data", "03-analysis_ready_datasets", "01-dataset_complete-MetaPhlAn.tse")

if(params$overwrite_TSE_out == "yes"){
  unlink(outputName, recursive = TRUE)
}

write_TSE_to_dir(MetaPhlAn_data_merged.tse, out.dir = outputName)
```

## HUMAnN 

```{r write complete HUMAnN dataset}
outputName <- file.path("~", "parkinsonMetaAnalysis_data", "03-analysis_ready_datasets", "01-dataset_complete-HUMAnN.tse")

if(params$overwrite_TSE_out == "yes"){
  unlink(outputName, recursive = TRUE)
}

write_TSE_to_dir(HUMAnN_data_merged.tse, out.dir = outputName)
```

# Session Info

```{r session info}
sessionInfo()
```
