---
title: "Parkinson Meta-Analysis with Random Forest"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  pMD_data_type: "relative_abundance"
  pseudocount: 1e-8
  transform: "asin_sqrt"
  prevalThresh: 0.05
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(mia)
library(magrittr)
library(tidyverse)
library(parkinsonsMetagenomicData)
library(ggtext)

# Random forest packages
library(ranger)
library(pROC)


library(ggpubr)
# for table visualizations
library(DT)

library(biobakeryUtils)

theme_set(theme_bw())

# create base directory output
outdir <- "results"
dir.create(outdir)
```

```{r, cache=TRUE}
input.path <- list.files("../Data", pattern = paste0(params$pMD_data_type, ".tse"), full.names = TRUE)
Input.tse <- read_TSE_from_dir(input.path)

# Define A Parkinson_1_0 variable. controls may have other diseases too
Input.tse@colData$Parkinson_1_0 <- factor(ifelse(grepl("Parkinson", Input.tse@colData$disease), 1, 0))

# convert NA in age_group to "unknown"
Input.tse@colData$age_group <- factor(ifelse(is.na(Input.tse@colData$age_group), "Unknown", Input.tse@colData$age_group))

# create number of diseases variables
Input.tse@colData$number_diseases <- ifelse(Input.tse@colData$disease == "Healthy", 0, str_count(Input.tse@colData$disease, ";") + 1)


# assay-wise curations
assay(Input.tse, params$transform) <- asin(sqrt(assay(Input.tse, params$pMD_data_type)))
# calculate CLR 
Input.tse <- transformAssay(Input.tse, assay.type = params$pMD_data_type, method = "clr", pseudocount = as.numeric(params$pseudocount))

Input.tse
```

# keep only datasets that are usable (enough case and controls)

```{r}
cases_ctrls_vs_study_name <- colData(Input.tse) %>% 
  as.data.frame() %>% 
  split.data.frame(.$study_name) %>% 
  lapply(function(x) as.data.frame(table(x$disease, useNA = "always"))) %>% 
  bind_rows(.id = "study_name")
  
cases_ctrls_vs_study_name_wider <- cases_ctrls_vs_study_name %>% 
  pivot_wider(names_from = Var1, values_from = Freq, values_fill = 0)

study_names_usable <- cases_ctrls_vs_study_name_wider$study_name[cases_ctrls_vs_study_name_wider$Healthy >= 10 & cases_ctrls_vs_study_name_wider$`Parkinson Disease` >= 10]

Input_usableDatasets.tse <- Input.tse[,Input.tse@colData$study_name %in% study_names_usable]
```

# Machine Learning (RF) of PD vs controls (leave one out) with `ranger`

  0 - Data were pre-normalized with the same approach 
  as the previous meta-analysis
  
  1 - Train microbiome on all vs one dataset
  
  2 - test on the left-out dataset
  
  3 - Repeat for all

Important: `ranger` doesn't seem to have a cross-validation step in the training.
SIAMCAT has it, and it is likely what is taking so long to complete.

```{r}
library(ranger)
library(pROC)

auroc.all <- tibble(
  study_left_out = character(0),
  AUROC = double(0),
  AUROC_P = double(0)
)

datasets <- unique(Input_usableDatasets.tse@colData$study_name)

#--------------------------------
feat_matrix <- assay(Input_usableDatasets.tse, params$transform)
metadata <- as.data.frame(colData(Input_usableDatasets.tse))
datasets <- unique(metadata$study_name)

t0_ranger <- Sys.time()

for(ds in datasets) {
  
  train_idx <- metadata$study_name != ds
  train_features <- t(feat_matrix[, train_idx])
  train_labels <- factor(metadata$Parkinson_1_0[train_idx])
  
  cat(sprintf("training on %s", ds))
  
  rf_model <- ranger(
    x = train_features,
    y = train_labels,
    num.trees = 1000,
    mtry = 10, 
    probability = TRUE,
    num.threads = 8,
    seed = 1234
  )
  
  for(ds2 in datasets) {
    
    test_idx <- metadata$study_name == ds2
    test_features <- t(feat_matrix[, test_idx])
    test_labels <- metadata$Parkinson_1_0[test_idx]
    
    cat(sprintf(" | testing on %s", ds2))
    
    pred_probs <- predict(rf_model, data = test_features)$predictions[, "1"]
    
    roc_obj <- roc(test_labels, pred_probs, quiet = TRUE)
    # random null distribtion for test
    roc2 <- roc(sample(0:1, size = length(test_labels), replace = TRUE), pred_probs, quiet = TRUE)
    # extract auc from the roc curve ()
    auc_value <- as.numeric(auc(roc_obj))

    # Test against null hypothesis (random labels, AUC = 0.5)
    # This creates a "null" ROC by comparing against random predictions
    p_value <- roc.test(roc_obj, roc2 = roc2, method = "delong")$p.value

    auroc.all <- add_row(auroc.all, 
      study_left_out = ds,
      AUROC = auc_value,
      AUROC_P = p_value
    )
  }
}

t1_ranger <- Sys.time()
```

```{r}
write_tsv(auroc.all, sprintf("%s/rf-PD_ctrl-LODO_%s.tsv", outdir, "ranger"))
datatable(auroc.all %>% filter(AUROC != 1))
```


# Same but with SIAMCAT

```{r}
library(SIAMCAT)

datasets <- unique(Input_usableDatasets.tse@colData$study_name)

auroc.all <- tibble(
  study_left_out = character(0),
  AUROC = double(0)
)

t0_siamcat <- Sys.time()

for(ds in datasets){
  
  # subset data to only the wanted dataset
  Input_subset.tse <- Input_usableDatasets.tse[, Input_usableDatasets.tse@colData$study_name != ds]
  
  # create SIAMCAT object
  sc.obj.train <- siamcat(feat=assay(Input_subset.tse, params$transform), meta=as.data.frame(colData(Input_subset.tse)), 
                          label='Parkinson_1_0', case=1)
  
  # normalize features
  sc.obj.train <- normalize.features(sc.obj.train, norm.method = 'pass', feature.type = "original")
  
  cat(sprintf("training on %s", ds))
  
  # Create data split
  sc.obj.train <- create.data.split(
    sc.obj.train,
    num.folds = 10,
    num.resample = 5,
    inseparable = NULL
  )
  
  # train Random Forest model
  sc.obj.train <- train.model(
    sc.obj.train,
    method = "randomForest",
    feature.type = "normalized",
    verbose = 3, 
    param.set = list(ntree = 1000, mtry = 10) # 1000 trees for training and 10 features per tree split (aka 'leaves')
  )
  
  # iteratively test all except 1 dataset into the left out dataset
  for(ds2 in datasets){
    # get only one dataset
    Input_subset2.tse <- Input_usableDatasets.tse[, Input_usableDatasets.tse@colData$study_name == ds]
    cat(sprintf(" | testing on %s", ds2))
    
  # create SIAMCAT object of test set
  sc.obj.test <- siamcat(
    feat = assay(Input_subset2.tse, params$transform),
    meta = as.data.frame(colData(Input_subset2.tse)),
    label = 'Parkinson_1_0',
    case = 1
  )
  # normalized test set (in our case it is pre-normalized)
  sc.obj.test <- normalize.features(sc.obj.test, norm.method = 'pass', feature.type = "original")
  # predict case/control status
  sc.obj.test <- make.predictions(sc.obj.train, siamcat.holdout = sc.obj.test)
  # evaluate predictions 
  sc.obj.test <- evaluate.predictions(sc.obj.test)
  
  # calculate AUROC p-value (can't calculate it in SIAMCAT yet)
  # roc2 <- roc.test(sc.obj.test@eval_data$roc, roc(sample(0:1, size = nrow(sc.obj.test@pred_matrix), replace = TRUE), pred_probs, quiet = TRUE)
  
  auroc.all <- auroc.all %>% 
                add_row(
                  study_left_out=ds, 
                  AUROC= as.double(eval_data(sc.obj.test)$auroc)
                  )
  }
}
t1_siamcat <- Sys.time()
```

```{r}
write_tsv(auroc.all, sprintf("%s/rf-PD_ctrl-LODO_%s.tsv", outdir, "SIAMCAT"))
datatable(auroc.all %>% filter(AUROC != 1))
```

# The quick ranger implementation makes little difference in our case but is much faster

```{r}
ranger_res <- read_tsv(sprintf("%s/rf-PD_ctrl-LODO_%s.tsv", outdir, "ranger"))
siamcat_res <- read_tsv(sprintf("%s/rf-PD_ctrl-LODO_%s.tsv", outdir, "SIAMCAT"))

runtime_ranger <- t1_ranger - t0_ranger
runtime_siamcat <- t1_siamcat - t0_siamcat

results_joined <- inner_join(ranger_res, siamcat_res, by = study_left_out, suffix = c(".ranger", ".siamcat"))
```

# Session Info

```{r}
sessionInfo()
```
