---
title: "Suppelementary Figures"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_tool: "MetaPhlAn"
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(mia)
library(magrittr)
library(tidyverse)
library(parkinsonsMetagenomicData)
library(ggtext)

library(ggpubr)
library(meta)

# for table visualizations
library(DT)
library(kableExtra)

library(biobakeryUtils)
theme_set(theme_bw())

# create base directory output
outdir <- sprintf("SupplementaryFigures")
dir.create(outdir, recursive = TRUE)

pd_color_palette <- c("Control" = "gray60", "PD" = "#D9565CFF")

# generate an incremental figure index
suppFigIndx <- 1
```

```{r prepare data for plot}
Input.tse <- read_TSE_from_dir(sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_complete-%s.tse", params$biobakery_tool))

# Parkinson as target condition
Input_parkinson.tse <- Input.tse[, Input.tse$target_condition == "Parkinson Disease"]
Input_parkinson.tse <- Input_parkinson.tse[rowSums(assay(Input_parkinson.tse)) > 0, ]

# drop unused level of the parkinson_for_analysis
Input_parkinson.tse$parkinson_for_analysis <- droplevels(Input_parkinson.tse$parkinson_for_analysis)

# Each dataset should have at least 30 samples in cases and 30 in controls
ok_datasets <- Input_parkinson.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  filter(
    sum(parkinson_for_analysis == "Control") > 10,
    sum(parkinson_for_analysis == "PD") > 10
  )

## filter TSE to restrict to 'ok' datasets 
Input_parkinson_usableDatasets.tse <- Input_parkinson.tse[,Input_parkinson.tse$study_name %in% unique(ok_datasets$study_name)]

# prepare data for plot "reads per dataset color by PD status"
data_for_plot.df <- as.data.frame(colData(Input_parkinson_usableDatasets.tse)) %>% 
  mutate(million_reads = reads_processed/1e6, study_name = factor(study_name))
```

## Supplementary Figure `r suppFigIndx` - Prevalence and detection filters

Two options are possible: one global and one per-dataset prevalence and abundance
filter.

The per-dataset prevalence results in fewer features, because some datasets have
very few samples. This would lead to an unfairly strict filter. 

```{r}
preval_curve.df <- data.frame(
  preval_thr = seq(0, 0.9, 0.005),
  nFeatures = sapply(seq(0, 0.9, 0.005), function(thr) {
    return(nrow(subsetByPrevalent(Input_parkinson_usableDatasets.tse, assay.type = "relative_abundance", prevalence = thr, detection = 0)))  
})
)

prevalence_curve.gg <- preval_curve.df %>% 
  ggplot(aes(x = preval_thr*100, y = nFeatures)) + 
  geom_point() + 
  geom_line(aes(group =1)) + 
  geom_vline(xintercept = 5, color = "red2") + 
  labs(
    title = "Prevalence curve flattens at 5% prevalence",
    subtitle = "Prevalence threshold of 5% was chosen",
    y = "N Features retained by filter",
    x = "Prevalence (%)"
  )

#--------------------------- same test but with detection

detect_curve.df <- data.frame(
  detection_thr = 10^-c(8:1),
  nFeatures = sapply(10^-c(8:1), function(thr) {
    return(nrow(subsetByPrevalent(Input_parkinson_usableDatasets.tse, assay.type = "relative_abundance", prevalence = 0, detection = thr)))  
})
)

detection_curve.gg <- detect_curve.df %>% 
  ggplot(aes(x = detection_thr*100, y = nFeatures)) + 
  geom_point() + 
  geom_line(aes(group =1)) + 
  scale_x_log10() + 
  geom_vline(xintercept = 1e-4, color = "red2") + 
  labs(
    title = "Detection threshold starts to matter at ~ 1%",
    subtitle = "Detection threshold of 1e-4 was chosen",
    y = "N Features retained by filter",
    x = "Detection minimum abundance ([0-1])"
  )

preval_detect_plots.gg <- ggarrange(prevalence_curve.gg, detection_curve.gg)


ggsave(plot = preval_detect_plots.gg, filename = sprintf("%s/SupplementaryFigure %s-Prevalence and detection thresholds.svg", outdir, suppFigIndx), height =4.5, width = 9)
saveRDS(preval_detect_plots.gg, file = sprintf("%s/SupplementaryFigure %s-Prevalence and detection thresholds.rds", outdir, suppFigIndx))

suppFigIndx <- suppFigIndx + 1
preval_detect_plots.gg
```

## Supplementary figure `r suppFigIndx` - read depth does not vary by PD-Control status

```{r calculate stats and make plot}
stats.df <- data_for_plot.df %>% 
  group_by(study_name) %>% 
  rstatix::t_test(million_reads ~ parkinson_for_analysis) %>% 
  ungroup() %>%                              # <-- ungroup BEFORE adjusting
  rstatix::adjust_pvalue(method = "BH") %>% 
  rstatix::add_xy_position(x = "parkinson_for_analysis") %>% 
  mutate(
    p.adj.fmt = format_p_value(p.adj, digits = 2)
  )

#' this test is a omnibus test that answers the question: 
#' "are there differences in read depth between PD and Controls once we account
#' for per-dataset differences in depth (by blocking for dataset)?"
random_eff_test <- lmerTest::lmer(million_reads ~ parkinson_for_analysis + (1 | study_name), data = data_for_plot.df)
random_eff_tidy.df <- broom.mixed::tidy(random_eff_test)
random_eff_params <- unlist(random_eff_tidy.df[random_eff_tidy.df$term == "parkinson_for_analysisPD", c("estimate", "p.value")])

signif_label <- sprintf("Eff. size = %.3f; P = %.3f", random_eff_params[1], random_eff_params[2])

reads_vs_dataset_color_PD <- ggplot(data_for_plot.df, aes(x = study_name, y = million_reads, color = parkinson_for_analysis)) +
  geom_boxplot() + 
  ggpubr::stat_pvalue_manual(data = stats.df, x = "study_name", label = "p.adj.fmt") + 
  scale_color_manual(values = pd_color_palette) + 
  annotate("text", x = -Inf, y = Inf, label = signif_label, hjust = -0.1, vjust = 1.5) +
  labs(
    color = NULL,
    x = NULL,
    y = "Million reads per sample"
  ) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), legend.position = "top")

# save plot
ggsave(plot = reads_vs_dataset_color_PD, filename = sprintf("%s/SupplementaryFigure %s-Read-depth-by-dataset-and-PD.svg", outdir, suppFigIndx), height = 6, width = 6)
saveRDS(reads_vs_dataset_color_PD, file = sprintf("%s/SupplementaryFigure %s-Read-depth-by-dataset-and-PD.rds", outdir, suppFigIndx))

suppFigIndx <- suppFigIndx + 1

```

# Session Info

```{r}
sessionInfo()
```

```