---
title: "Parkinson Meta-Analysis"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  pMD_data_type: "relative_abundance"
  DA_method: "limma"
---

Choices for `relative_abundance` data:
  - prevalence and abundance filter
  - limma-voom on simil-counts data (relative abundance x reads processed)
  - no standardization after modeling
  - meta-analyze eff. size and standard errors of the modeling estimates

# Data and environment preparation

```{r setup, message=FALSE, warning=FALSE}
# ------ prepare outdir --------
outdir <- sprintf("results/%s/%s", params$pMD_data_type, params$DA_method)
dir.create(outdir, recursive = TRUE)
# ------------------------------

library(meta)
library(limma)
library(biobakeryUtils)
library(mia)
library(openxlsx2)
library(tidyverse)
library(knitr)
Input_parkinson_usableDatasets.tse <- read_TSE_from_dir(sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_PD-%s.tse/", params$pMD_data_type))

Input_parkinson_usableDatasets.tse
```

## Case of `limma`

```{r, eval=(params$DA_method == "limma")}
assay(Input_parkinson_usableDatasets.tse, "counts") <- sweep(assay(Input_parkinson_usableDatasets.tse), MARGIN = 2, STATS = Input_parkinson_usableDatasets.tse@colData$reads_processed, FUN = "*")

per_dataset_DA_results.tb <- Input_parkinson_usableDatasets.tse %>% 
  splitOn(group = "study_name") %>% 
  lapply(function(x.tse) {
    x.tse <- x.tse[rowSums(assay(x.tse, "relative_abundance")) > 0,]
    return(limmaTSE(tse = x.tse, assay.type = "counts", voom.transform = TRUE, formula = ~ parkinson_for_analysis, coef = "parkinson_for_analysis1", p.adj_method = "BH"))
  }
  ) %>% bind_rows(.id = "study_name")
```

## Case of `Maaslin2`

```{r, eval=(params$DA_method == "maaslin2")}
per_dataset_DA_results.tb <- Input_parkinson_usableDatasets.tse %>%
  splitOn(group = "study_name") %>%
  lapply(function(x.tse) {
    x.tse <- x.tse[rowSums(assay(x.tse, "relative_abundance")) > 0, ]
    return(
      maaslin2TSE(
        tse = x.tse,
        assay.type = "relative_abundance",
        outdir = outdir,
        outdir_delete = TRUE,
        normalization = "NONE",
        min_prevalence = 0,
        min_abundance = 0,
        plot_heatmap = FALSE,
        plot_scatter = FALSE,
        save_scatter = FALSE,
        standardize = FALSE,
        min_variance = 0,
        transform = "LOG",
        max_significance = 0.1,
        fixed_effects = "parkinson_for_analysis",
        correction = "BH"
      )
    )
  }) %>% bind_rows(.id = "study_name")
```

## meta-analyze datasets (`REML`)

```{r, meta-analysis functions definition}
MetaAnalysis_results <- metaAnalyze(res.df = per_dataset_DA_results.tb, grouping.col = "study_name", beta.col = "Beta", se.col = "SE", method = "REML", test = "knha", minStudies.n = 6)
```

## Results at a glance

```{r}
feats_tot <- nrow(MetaAnalysis_results$MetaAnalysis)

heterog <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(Het.Pval < 0.1) %>% 
  nrow()

notSignif <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(
    Het.Pval >= 0.1, 
    P.Value.adj >+ 0.1
    ) %>% 
  nrow()

Positive_assoc <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(
    Het.Pval >= 0.1, 
    P.Value.adj < 0.1,
    Beta.pool > 0
    ) %>% 
  nrow()

Negative_assoc <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(
    Het.Pval >= 0.1, 
    P.Value.adj < 0.1,
    Beta.pool < 0
    ) %>% 
  nrow()

cbind(
  c("Tot tested", "N. heterogeneous", "Not Signif. (FDR >= 0.1)", "Signif. Positive assoc.", "Signif. Negative assoc."),
  c(feats_tot, heterog, notSignif, Positive_assoc, Negative_assoc)
) %>% kable()
```

# Write results

## Per-dataset results as long format

```{r}
write_tsv(per_dataset_DA_results.tb, file = sprintf("%s/01-singleDataset_DA.tsv", outdir))
```

## Meta-analysis results (as `.xlsx`)

```{r}
# meta-analysis results as list of Cohen D, SE(CohenD)
openxlsx2::write_xlsx(MetaAnalysis_results, file = sprintf("%s/02-metaAnalysis.xlsx", outdir))
```

# Session Info

```{r}
sessionInfo()
```
