---
title: "Parkinson Meta-Analysis"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_tool: "MetaPhlAn"
---

Choices for `relative_abundance` data:
  - prevalence and abundance filter
  - limma-voom on simil-counts data (relative abundance x reads processed)
  - no standardization after modeling
  - meta-analyze eff. size and standard errors of the modeling estimates

# Data and environment preparation

```{r setup, message=FALSE, warning=FALSE}
# ------ prepare outdir --------
outdir <- sprintf("results/%s", params$biobakery_tool)
dir.create(outdir, recursive = TRUE)
# ------------------------------

library(meta)
library(limma)
library(biobakeryUtils)
library(mia)
library(openxlsx2)
library(tidyverse)
library(knitr)
Input_parkinson_usableDatasets.tse <- read_TSE_from_dir(sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/02-dataset_PD-%s.tse/", params$biobakery_tool))

Input_parkinson_usableDatasets.tse
```

## Single dataset analysis with `limma-voom`

Important featueres of `voom`: 
  - variance stabilizing
  - uses pseudocount of 0.5
  - transforms pseudocount-added samples with counts per million and log to
  - formula is, for sample $x$: 
  
  $\log_2{\frac{x_p}{\sum_i^n{x_i} + 1}\times10^6}$
  
```{r limma regression}
per_dataset_DA_results_limma.tb <- Input_parkinson_usableDatasets.tse %>% 
  splitOn(group = "study_name") %>% 
  lapply(function(x.tse) {
    x.tse <- x.tse[rowSums(assay(x.tse)) > 0,]
    return(limmaTSE(tse = x.tse, assay.type = "counts", voom.transform = TRUE, formula = ~ parkinson_for_analysis, coef = "parkinson_for_analysis1", p.adj_method = "BH"))
  }
  ) %>% bind_rows(.id = "study_name")
```

## Single dataset analysis with `Maaslin2`

```{r maaslin2 regression, include=FALSE}
per_dataset_DA_results_maaslin2.tb <- Input_parkinson_usableDatasets.tse %>%
  splitOn(group = "study_name") %>%
  lapply(function(x.tse) {
    x.tse <- x.tse[rowSums(assay(x.tse, "relative_abundance")) > 0, ]
    return(
      maaslin2TSE(
        tse = x.tse,
        assay.type = "relative_abundance",
        outdir = tempfile(pattern = "maaslin2_"),
        min_variance = 0,
        transform = "LOG",
        max_significance = 0.1,
        fixed_effects = "parkinson_for_analysis",
        correction = "BH",
        outdir_delete = TRUE,
        normalization = "NONE",
        min_prevalence = 0,
        min_abundance = 0,
        plot_heatmap = FALSE,
        plot_scatter = FALSE,
        save_scatter = FALSE,
        standardize = FALSE
      )
    )
  }) %>% bind_rows(.id = "study_name")
```

## meta-analyze datasets (`REML`)

```{r, meta-analysis limma}
MetaAnalysis_results_limma <- metaAnalyze(res.df = per_dataset_DA_results_limma.tb, grouping.col = "study_name", beta.col = "Beta", se.col = "SE", method = "REML", test = "knha", minStudies.n = 6)
```


```{r, meta-analysis maaslin2}
MetaAnalysis_results_maaslin2 <- metaAnalyze(res.df = per_dataset_DA_results_maaslin2.tb, grouping.col = "study_name", beta.col = "Beta", se.col = "SE", method = "REML", test = "knha", minStudies.n = 6)
```

## Results table at a glance

```{r results table for both method meta-analyses}
metaAnalysis_glance <- function(MetaAnalysis_results, signif_thr = 0.1, het_signif_thr = 0.1){
  feats_tot <- nrow(MetaAnalysis_results$MetaAnalysis)

heterog <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(Het.Pval < het_signif_thr) %>% 
  nrow()

notSignif <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(
    Het.Pval >= het_signif_thr, 
    P.Value.adj >= signif_thr
    ) %>% 
  nrow()

Positive_assoc <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(
    Het.Pval >= het_signif_thr, 
    P.Value.adj < signif_thr,
    Beta.pool > 0
    ) %>% 
  nrow()

Negative_assoc <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(
    Het.Pval >= het_signif_thr, 
    P.Value.adj < signif_thr,
    Beta.pool < 0
    ) %>% 
  nrow()
  
  out.df <- data.frame(
    result_name = c("Tot Tested", "Heterogeneous", "Not Significant", "Positive association", "Negative association"),
    result_condition = c("N rows", sprintf("Het.Pval >= %s", het_signif_thr),  sprintf("Het.Pval >= %s; P.adj >= %s", het_signif_thr, signif_thr), sprintf("Het.Pval >= %s; P.adj < %s; Beta > 0", het_signif_thr, signif_thr), sprintf("Het.Pval >= %s; P.adj < %s; Beta < 0", het_signif_thr, signif_thr)),
    n_features = c(feats_tot, heterog, notSignif, Positive_assoc, Negative_assoc))

  
  return(out.df)
}

meta_analyses_at_a_glance <- list(
  "limma" = MetaAnalysis_results_limma,
  "maaslin2" = MetaAnalysis_results_maaslin2
) %>% 
  lapply(metaAnalysis_glance) %>% 
  reduce(full_join, by = c("result_name", "result_condition"),suffix = c("_limma", "_maaslin2"))

kable(meta_analyses_at_a_glance)
```

# Write results

## limma

### Full differential abundance results

```{r write per dataset limma regressions in long format}
write_tsv(per_dataset_DA_results_limma.tb, file = sprintf("%s/01-%s_per_dataset_DA_results_limma.tsv", outdir, params$biobakery_tool))
```

### Meta-analysis results (as `.xlsx`)

```{r write beta SE matrices and metaAnalysis of limma}
# meta-analysis results as list of Cohen D, SE(CohenD)
openxlsx2::write_xlsx(MetaAnalysis_results_limma, file = sprintf("%s/02-%s_MetaAnalysis_results_limma.xlsx", outdir, params$biobakery_tool))
```

## maaslin2

### Full differential abundance results

```{r write per dataset maaslin2 regressions in long format}
write_tsv(per_dataset_DA_results_maaslin2.tb, file = sprintf("%s/01-%s_per_dataset_DA_results_maaslin2.tsv", outdir, params$biobakery_tool))
```

### Meta-analysis results (as `.xlsx`)

```{r write beta SE matrices and metaAnalysis of maaslin2}
# meta-analysis results as list of Cohen D, SE(CohenD)
openxlsx2::write_xlsx(MetaAnalysis_results_maaslin2, file = sprintf("%s/02-%s_MetaAnalysis_results_maaslin2.xlsx", outdir, params$biobakery_tool))
```

# Session Info

```{r session info}
sessionInfo()
```
