---
title: "Study tables, MetaPhlAn alpha and beta diversities"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_tool: "MetaPhlAn"
---

```{r}
# ------ prepare outdir --------
outdir <- file.path("results", "01-stats-alpha-beta", params$biobakery_tool)
dir.create(outdir, recursive = TRUE)
# ------------------------------
library(jsonlite)
library(knitr)
library(gtsummary)
library(ggtext)
library(ggsci)
library(mia)
library(tidyverse)
library(biobakeryUtils)
library(ChocoPhlAnTaxonomies)

Input.tse <- read_TSE_from_dir(sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_complete-%s.tse", params$biobakery_tool))

Input.tse
```

# `parkinsonsMetagenomicData` kept after each filtering step

```{r}
metaphlan_filters <- read_json("~/parkinsonMetaAnalysis_data/01-pMD_datasets/00-pMD_MetaPhlAn_sample_filtering_steps.json")
humann_filters <- read_json("~/parkinsonMetaAnalysis_data/01-pMD_datasets/00-pMD_HUMAnN_sample_filtering_steps.json")

samples_kept_after_filtering.df <- tibble(
  filterCondition = names(metaphlan_filters),
  MetaPhlAn = sapply(metaphlan_filters, length),
  HUMAnN = sapply(humann_filters, length)
)

kable(samples_kept_after_filtering.df)
```

# Table 1 - dataset x PD, unavailable covariates limit meta-analysis strength

## All samples available after filtering

```{r table of samples per condition}
table1_complete <- Input.tse %>% 
  colData() %>% 
  as.data.frame() %>% 
  select(`Study` = study_name, Country, `Age Gr.` = age_group, Sex = sex, `N. diseases` = number_diseases, parkinson_for_analysis) %>% 
   tbl_summary(
    by = parkinson_for_analysis, # Optional: stratify by a variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing_text = "Missing"
  ) %>%
  add_overall() %>% # Add an overall column
  modify_header(label ~ "**Characteristic**") # Customize header

table1_complete
```

## Datasets with available age and sex information

```{r table of samples per condition with both sex and age_group available}
table1_availSex <- Input.tse %>% 
  colData() %>% 
  as.data.frame() %>% 
  filter(complete.cases(select(., age_group, sex))) %>% 
  select(`Study` = study_name, Country, `Age Gr.` = age_group, Sex = sex, `N. diseases` = number_diseases, parkinson_for_analysis) %>% 
   tbl_summary(
    by = parkinson_for_analysis, # Optional: stratify by a variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing_text = "Missing"
  ) %>%
  add_overall() %>% # Add an overall column
  modify_header(label ~ "**Characteristic**") # Customize header
```

## Datasets with at least 10 individuals in PD and Control groups

```{r filter dataset for parkinson meta-analysis}
# Parkinson as target condition
Input_parkinson.tse <- Input.tse[, Input.tse$target_condition == "Parkinson Disease"]
Input_parkinson.tse <- Input_parkinson.tse[rowSums(assay(Input_parkinson.tse)) > 0, ]

# drop unused level of the parkinson_for_analysis
Input_parkinson.tse$parkinson_for_analysis <- droplevels(Input_parkinson.tse$parkinson_for_analysis)

# Each dataset should have at least 30 samples in cases and 30 in controls
ok_datasets <- Input_parkinson.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  filter(
    sum(parkinson_for_analysis == "Control") > 10,
    sum(parkinson_for_analysis == "PD") > 10
  )

## filter TSE to restrict to 'ok' datasets 
Input_parkinson_usableDatasets.tse <- Input_parkinson.tse[,Input_parkinson.tse$study_name %in% unique(ok_datasets$study_name)]

# ## Prevalence and detection filter
# Input_parkinson_usableDatasets.tse <- subsetByPrevalent(Input_parkinson_usableDatasets.tse, prevalence = params$preval_thresh, detection = params$detect_thresh)

Input_parkinson_usableDatasets.tse
```

```{r table of samples per condition with sufficient case/ctrl samples}

table1_ok_Parkinson <- Input_parkinson_usableDatasets.tse %>% 
  colData() %>% 
  as.data.frame() %>% 
  filter(complete.cases(select(., age_group, sex))) %>% 
  select(`Study` = study_name, Country, `Age Gr.` = age_group, Sex = sex, `N. diseases` = number_diseases, parkinson_for_analysis) %>% 
   tbl_summary(
    by = parkinson_for_analysis, # Optional: stratify by a variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing_text = "Missing"
  ) %>%
  add_overall() %>% # Add an overall column
  modify_header(label ~ "**Characteristic**") # Customize header

table1_ok_Parkinson
```

# Alpha diversities

```{r calculate alpha diverity metrics}
if(params$biobakery_tool == "MetaPhlAn"){
  alphas <- c("observed_richness", "shannon_diversity", "faith")
  Input_parkinson_usableDatasets.tse <- AddPhyloTree_to_mpa_tse(Input_parkinson_usableDatasets.tse, chocophlan_datestamp = "202403")
} else {
  alphas <- c("observed_richness", "shannon_diversity")  
}

# Calculate alphas
Input_parkinson_usableDatasets.tse <- addAlpha(Input_parkinson_usableDatasets.tse, assay.type = "relative_abundance", index = alphas)
```

## PCoA - bray curtis

```{r MDS params}
#--- variable setup
beta.method <- "bray"
beta.name <- "bray"
```


```{r MDS plots}
Input_parkinson_usableDatasets.tse <- addMDS(Input_parkinson_usableDatasets.tse, assay.type = "relative_abundance", method = beta.method, name = beta.name)

variances_explained <- attr(reducedDim(Input_parkinson_usableDatasets.tse, beta.name), "eig")
prop_variances_explained <- variances_explained/sum(variances_explained)

betadiv_data.df <- cbind.data.frame(as.data.frame(colData(Input_parkinson_usableDatasets.tse)), reducedDim(Input_parkinson_usableDatasets.tse, beta.name) %>% as.data.frame() %>% set_names(paste("MDS", 1:ncol(.))))

colorVars <- c("Country", "study_name", "parkinson_for_analysis")

PCoA_plots.list <- map(colorVars, function(colorVar) {
  betadiv_data.df %>%
    ggplot(aes(x = `MDS 1`, y = `MDS 2`, color = !!sym(colorVar), fill = !!sym(colorVar))) + 
  geom_point(size = 2.5, stroke = 1, color = "black") +
  geom_point(size = 2.5) +
  stat_ellipse(show.legend = FALSE) + 
  ggsci::scale_color_jco() +
  ggsci::scale_fill_jco() +
  theme(legend.position = "top", legend.text = element_markdown()) +
  labs(color = colorVar,
       fill = colorVar,
       x = paste0("PCoA Axis 1 [", round(prop_variances_explained[1]*100, 1), "%]"),
       y = paste0("PCoA Axis 2 [", round(prop_variances_explained[2]*100, 2), "%]")
  )
}) %>% 
  set_names(sprintf("MDS_%s_color_%s", beta.name, colorVars))

PCoA_plots.list[[3]]
```

### PERMANOVA

$$
Bray-Curtis \sim study~name + parkinson + reads~processed
$$

```{r PERMANOVA, cache=TRUE}
PERMANOVA_name <- paste(beta.name, "PERMANOVA", sep = "_")

Input_parkinson_usableDatasets.tse <- addPERMANOVA(
  Input_parkinson_usableDatasets.tse,
  assay.type = "relative_abundance",
  method = beta.method,
  name = PERMANOVA_name,
  formula = x ~ study_name + parkinson_for_analysis + reads_processed,
  by = "margin",
  permutations = 2000, 
  parallel = 8
)

bray_permanova <- PERMANOVA_to_table(Input_parkinson_usableDatasets.tse, name = PERMANOVA_name)
```

# Output Figures (as `.rds`)

```{r figures outdir setup}
outdir_figures <- file.path(outdir, "Figures")
dir.create(outdir_figures)
```

## List of PCoA plots colored by different colData columns

```{r save MDS plots as .rds}
imap(PCoA_plots.list, function(plot.gg, pltName){
  saveRDS(plot.gg, file = sprintf("%s/01-%s.rds", outdir_figures, pltName))
})
```

# Output Tables (multiple formats)

```{r tables outdir setup}
outdir_tables <- file.path(outdir, "Tables")
dir.create(outdir_tables)
```

## Filtering steps - samples and features excluded (as `.tsv`)

```{r steps of samples and features exclusions}
write_tsv(rbind(samples_kept_after_filtering.df, c("Enough Cases and Controls per dataset", dim(Input_parkinson_usableDatasets.tse))), file = sprintf("%s/01-filtering-samples-and-features-excluded.tsv", outdir_tables))
```

## Three versions of study tables - overview of datasets and covariates (as `.docx`)

```{r write demographic tables}
as_flex_table(table1_complete) %>% flextable::save_as_docx(path = sprintf("%s/02-table1-complete.docx", outdir_tables))

as_flex_table(table1_availSex) %>% flextable::save_as_docx(path = sprintf("%s/02-table1-availSex.docx", outdir_tables))

as_flex_table(table1_ok_Parkinson) %>% flextable::save_as_docx(path = sprintf("%s/02-table1-okParkinson.docx", outdir_tables))
```

## Alpha diversity tables for future figures and modeling

```{r write alpha diversity data for future plots and stats}
data_for_alpha_calculations <- colData(Input_parkinson_usableDatasets.tse) %>% 
  as.data.frame() %>% 
  select(all_of(alphas), parkinson_for_analysis, age_group, number_diseases, reads_processed, study_name) %>% 
write_tsv(file = sprintf("%s/03-alpha_diversity_data.tsv", outdir_tables))
```

## Alpha diversity modeling results

```{r lm of alpha vs PD and other covariates}
# Model alpha diversity including covariates
alpha_regression_pd_study.lm.list <- sapply(alphas, function(x) {
  lm(as.formula(sprintf("%s ~ parkinson_for_analysis + study_name + number_diseases", x)), data = data_for_alpha_calculations)
}, simplify = FALSE) %>% 
  lapply(broom::tidy) %>%
  lapply(mutate, FDR = p.adjust(p.value, "BH"))

bind_rows(alpha_regression_pd_study.lm.list, .id = "alpha_metric") %>% 
  write_tsv(file = sprintf("%s/04-alpha_diversity_models_joined.tsv", outdir_tables))
```

## PERMANOVA table

```{r PERMANOVA calculation based on beta-diversity}
write_tsv(bray_permanova, file = sprintf("%s/05-%s.tsv", outdir_tables, PERMANOVA_name))
```

# Session Info

```{r}
sessionInfo()
```
