---
title: "Parkinson Meta-Analysis"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_tool: "MetaPhlAn"
  preval_thresh: 0.05
  detect_thresh: 1e-5
---

Choices for `relative_abundance` data:
  - prevalence and abundance filter
  - limma-voom on simil-counts data (relative abundance x reads processed)
  - no standardization after modeling
  - meta-analyze eff. size and standard errors of the modeling estimates

# Data and environment preparation

```{r setup, message=FALSE, warning=FALSE}
# ------ prepare outdir --------
outdir <- file.path("results", "02-PD-meta-analysis", params$biobakery_tool)
dir.create(outdir, recursive = TRUE)
# ------------------------------

library(meta)
library(limma)
library(biobakeryUtils)
library(mia)
library(openxlsx2)
library(tidyverse)
library(knitr)
library(VennDetail)

Input.tse <- read_TSE_from_dir(sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_complete-%s.tse", params$biobakery_tool))

Input.tse
```

```{r filter dataset for parkinson meta-analysis}
# Parkinson as target condition
Input_parkinson.tse <- Input.tse[, Input.tse$target_condition == "Parkinson Disease"]
Input_parkinson.tse <- Input_parkinson.tse[rowSums(assay(Input_parkinson.tse)) > 0, ]

# drop unused level of the parkinson_for_analysis
Input_parkinson.tse$parkinson_for_analysis <- droplevels(Input_parkinson.tse$parkinson_for_analysis)

# Each dataset should have at least 30 samples in cases and 30 in controls
ok_datasets <- Input_parkinson.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  filter(
    sum(parkinson_for_analysis == "Control") > 10,
    sum(parkinson_for_analysis == "PD") > 10
  )

## filter TSE to restrict to 'ok' datasets 
Input_parkinson_usableDatasets.tse <- Input_parkinson.tse[,Input_parkinson.tse$study_name %in% unique(ok_datasets$study_name)]

# ## Prevalence and detection filter
# Input_parkinson_usableDatasets.tse <- subsetByPrevalent(Input_parkinson_usableDatasets.tse, prevalence = params$preval_thresh, detection = params$detect_thresh)

Input_parkinson_usableDatasets.tse
```

```{r}
kable(table(Input_parkinson_usableDatasets.tse$study_name, Input_parkinson_usableDatasets.tse$parkinson_for_analysis, useNA = "always"))
```

## Single dataset analysis with `limma-voom`

Important featueres of `voom`: 
  - variance stabilizing
  - uses pseudocount of 0.5
  - transforms pseudocount-added samples with counts per million and log to
  - formula is, for sample $x$: 
  
  $\log_2{\frac{x_p + 0.5}{\sum_i^n{x_i} + 1}\times10^6}$
  
```{r limma regression}
per_dataset_DA_results_limma.tb <- Input_parkinson_usableDatasets.tse %>% 
  splitOn(group = "study_name") %>% 
  lapply(function(x.tse) {
    x.tse <- x.tse[rowSums(assay(x.tse)) > 0,]
    return(limmaTSE(tse = x.tse, assay.type = "counts", voom.transform = TRUE, formula = ~ parkinson_for_analysis, coef = "parkinson_for_analysisPD", p.adj_method = "BH"))
  }
  ) %>% bind_rows(.id = "study_name")
```

## Single dataset analysis with `Maaslin2`

```{r maaslin2 regression, include=FALSE}
per_dataset_DA_results_maaslin2.tb <- Input_parkinson_usableDatasets.tse %>%
  splitOn(group = "study_name") %>%
  lapply(function(x.tse) {
    x.tse <- x.tse[rowSums(assay(x.tse, "relative_abundance")) > 0, ]
    return(
      maaslin2TSE(
        tse = x.tse,
        assay.type = "relative_abundance",
        outdir = tempfile(pattern = "maaslin2_"),
        outdir_delete = TRUE,
        min_variance = 0,
        transform = "LOG",
        max_significance = 0.2,
        fixed_effects = "parkinson_for_analysis",
        correction = "BH",
        normalization = "NONE",
        min_prevalence = 0,
        min_abundance = 0,
        plot_heatmap = FALSE,
        plot_scatter = FALSE,
        save_scatter = FALSE,
        standardize = FALSE
      )
    )
  }) %>% bind_rows(.id = "study_name") %>% 
  filter(ExposureValue == "parkinson_for_analysisPD") %>% 
  # readjust the p-value because we are interested only in one comparison
  group_by(study_name) %>% 
  mutate(P.Value.adj = p.adjust(P.Value, "BH")) %>% 
  ungroup()
```

## meta-analyze results (`REML`)

```{r, meta-analysis limma}
MetaAnalysis_results_limma <- metaAnalyze(res.df = per_dataset_DA_results_limma.tb, grouping.col = "study_name", beta.col = "Beta", se.col = "SE", method = "REML", test = "knha")
```

```{r, meta-analysis maaslin2}
MetaAnalysis_results_maaslin2 <- metaAnalyze(res.df = per_dataset_DA_results_maaslin2.tb, grouping.col = "study_name", beta.col = "Beta", se.col = "SE", method = "REML", test = "knha")
```

## Results table at a glance {.tabset}

```{r results table for both method meta-analyses}
metaAnalysis_glance <- function(MetaAnalysis_results, signif_thr = 0.2, het_signif_thr = 0.1){
  
  featureName <- colnames(MetaAnalysis_results$MetaAnalysis)[grep("StudiesTested", colnames(MetaAnalysis_results$MetaAnalysis)) - 1]
  
  tot_tested <- MetaAnalysis_results$MetaAnalysis[[featureName]]
  
  heterog <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(Het.Pval < het_signif_thr) %>% 
  .[[featureName]]

notSignif <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(
    P.Value.adj >= signif_thr
    ) %>% 
  .[[featureName]]

Positive_assoc <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(
    P.Value.adj < signif_thr,
    Beta.pool > 0
    ) %>% 
  .[[featureName]]

Negative_assoc <- MetaAnalysis_results$MetaAnalysis %>% 
  filter(
    P.Value.adj < signif_thr,
    Beta.pool < 0
    ) %>% 
  .[[featureName]]
  

return(
  list(
    # "Tot Tested" = tot_tested,
    "Heterogeneous" = heterog,
    "Not Significant" = notSignif,
    "Positive" = Positive_assoc,
    "Negative" = Negative_assoc
  )
)
}

meta_analyses_at_a_glance <- list(
  "limma" = MetaAnalysis_results_limma,
  "maaslin2" = MetaAnalysis_results_maaslin2
) %>% 
  lapply(metaAnalysis_glance)
```

### limma

```{r upset plot of limma meta-analysis results}
meta_analyses_at_a_glance$limma %>% 
  upset_plot
```

### maaslin2

```{r upset plot of maaslin2 meta-analysis results}
meta_analyses_at_a_glance$maaslin2 %>% 
  upset_plot
```

## {-}

# Write results

## limma

### Full differential abundance results

```{r write per dataset limma regressions in long format}
write_tsv(per_dataset_DA_results_limma.tb, file = sprintf("%s/01-%s_per_dataset_DA_results_limma.tsv", outdir, params$biobakery_tool))
```

### Meta-analysis results (as `.xlsx`)

```{r write beta SE matrices and metaAnalysis of limma}
# meta-analysis results as list of Cohen D, SE(CohenD)
openxlsx2::write_xlsx(MetaAnalysis_results_limma, file = sprintf("%s/02-%s_MetaAnalysis_results_limma.xlsx", outdir, params$biobakery_tool))
```

## maaslin2

### Full differential abundance results

```{r write per dataset maaslin2 regressions in long format}
write_tsv(per_dataset_DA_results_maaslin2.tb, file = sprintf("%s/01-%s_per_dataset_DA_results_maaslin2.tsv", outdir, params$biobakery_tool))
```

### Meta-analysis results (as `.xlsx`)

```{r write beta SE matrices and metaAnalysis of maaslin2}
# meta-analysis results as list of Cohen D, SE(CohenD)
openxlsx2::write_xlsx(MetaAnalysis_results_maaslin2, file = sprintf("%s/02-%s_MetaAnalysis_results_maaslin2.xlsx", outdir, params$biobakery_tool))
```

# Session Info

```{r session info}
sessionInfo()
```
