---
title: "Parkinson Meta-Analysis with Random Forest"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_tool: "MetaPhlAn"
  assay_to_test: "relative_abundance"
  preval_thresh: 0.05
  detect_thresh: 1e-5
---

```{r setup, message=FALSE, warning=FALSE}

# create base directory output
outdir <- file.path("results", "03-RF-LODO-PD-vs-Control", params$biobakery_tool)
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

library(knitr)
opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(mia)
library(magrittr)
library(tidyverse)

# Random forest packages
library(ranger)
library(pROC)
library(SIAMCAT)

# for table visualizations
library(DT)

# for loading tree summarized experiment
library(biobakeryUtils)

# for summary plots
library(ggpubr) 
theme_set(theme_bw())

Input.tse <- read_TSE_from_dir(sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_complete-%s.tse", params$biobakery_tool))

# transform data with limma's method, which should be uniform for both MetaPhlAn and HUMAnN data
if(params$assay_to_test == "voom"){
assay(Input.tse, params$assay_to_test) <- apply(assay(Input.tse, "counts"), 2, function(x) log2(x + 0.5) - log2(sum(x + 1)) + log2(1e6))
}

Input.tse
```

```{r filter dataset for parkinson meta-analysis}
# Parkinson as target condition
Input_parkinson.tse <- Input.tse[, Input.tse$target_condition == "Parkinson Disease"]
Input_parkinson.tse <- Input_parkinson.tse[rowSums(assay(Input_parkinson.tse)) > 0, ]

# drop unused level of the parkinson_for_analysis
Input_parkinson.tse$parkinson_for_analysis <- droplevels(Input_parkinson.tse$parkinson_for_analysis)

# Each dataset should have at least 30 samples in cases and 30 in controls
ok_datasets <- Input_parkinson.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  filter(
    sum(parkinson_for_analysis == "Control") > 10,
    sum(parkinson_for_analysis == "PD") > 10
  )

## filter TSE to restrict to 'ok' datasets 
Input_parkinson_usableDatasets.tse <- Input_parkinson.tse[,Input_parkinson.tse$study_name %in% unique(ok_datasets$study_name)]

# ## Prevalence and detection filter
# Input_parkinson_usableDatasets.tse <- subsetByPrevalent(Input_parkinson_usableDatasets.tse, prevalence = params$preval_thresh, detection = params$detect_thresh)

Input_parkinson_usableDatasets.tse
```

```{r}
kable(table(Input_parkinson_usableDatasets.tse$study_name, Input_parkinson_usableDatasets.tse$parkinson_for_analysis, useNA = "always"))
```

# Machine Learning (RF) of PD vs controls (leave one out) with `ranger`

  0 - Data were pre-normalized with the same approach 
  as the previous meta-analysis
  
  1 - Train microbiome on all vs one dataset
  
  2 - test on the left-out dataset
  
  3 - Repeat for all

Important: `ranger` doesn't seem to have a cross-validation step in the training.
SIAMCAT has it, and it is likely what is taking so long to complete.

```{r rf classification with ranger}

feat_matrix <- assay(Input_parkinson_usableDatasets.tse, params$assay_to_test)
metadata    <- as.data.frame(colData(Input_parkinson_usableDatasets.tse))
datasets    <- unique(metadata$study_name)

t0_train_ranger <- Sys.time()

trained_models_ranger.list <- sapply(datasets, function(ds) {
    # prepare data for training
    train_idx      <- metadata$study_name != ds
    train_features <- t(feat_matrix[, train_idx])
    train_labels   <- factor(metadata$parkinson_for_analysis[train_idx])
    
    # train rf model on all datasets except ds
    rf_model <- ranger(
      x           = train_features,
      y           = train_labels,
      num.trees   = 2000,
      mtry        = 10,
      probability = TRUE,
      num.threads = 6,
      seed        = 1234
    )
    
    return(rf_model)
    }, 
    simplify = FALSE, USE.NAMES = TRUE
    )

t1_train_ranger <- Sys.time()

# test each model on its dataset
t0_test_ranger <- Sys.time()

tested_models_ranger.list <- imap(trained_models_ranger.list, function(rf_model, ds){
  # create test sets
  test_idx      <- metadata$study_name == ds
  test_features <- t(feat_matrix[, test_idx])
  
  # true binary labels: 1 = PD, 0 = Control (mirrors SIAMCAT conversion)
  true_labels_binary <- as.integer(
    metadata$parkinson_for_analysis[test_idx] == "PD"
  )
  
  # get prediction probabilities
  pred_probs <- predict(rf_model, data = test_features)$predictions[, "PD"]
  
  # DeLong 95% Confidence intervals of AUC/AUROC
  roc_obj <- roc(response  = true_labels_binary,
                 predictor = pred_probs,
                 direction = "<",
                 quiet     = TRUE)
  ci_obj  <- ci.auc(roc_obj, method = "delong")
  
  # Wilcoxon p-value (same approach as SIAMCAT chunk)
  pd_scores      <- pred_probs[true_labels_binary == 1]
  control_scores <- pred_probs[true_labels_binary == 0]
  wilcox_res     <- wilcox.test(pd_scores, control_scores, alternative = "greater")
  
  # return 1-row data frame that is easy to bind together
  return(
    data.frame(
      AUROC = as.double(auc(roc_obj)),
      AUROC_lowerCI  = as.double(ci_obj[1]),
      AUROC_upperCI  = as.double(ci_obj[3]),
      p_value_wilcox = wilcox_res$p.value))
})
t1_test_ranger <- Sys.time()

rf_results_ranger.df <- bind_rows(tested_models_ranger.list, .id = "TestDataset")
```

```{r save ranger results}
# save all training models as .rds
saveRDS(trained_models_ranger.list, file = file.path(outdir, sprintf("rf_training-PD_ctrl-LODO_%s_%s.rds", "ranger", params$assay_to_test)))
# save all polished results
write_tsv(rf_results_ranger.df, file.path(outdir, sprintf("rf_results-PD_ctrl-LODO_%s_%s.tsv", "ranger", params$assay_to_test)))

datatable(rf_results_ranger.df)
```

# Machine Learning (RF) of PD vs controls (leave one out) with `SIAMCAT`

```{r rf classification with SIAMCAT}
library(future)
library(furrr)

plan(multisession, workers = 10)

# train with cross-validation 

t0_train_siamcat <- Sys.time()

trained_models_siamcat.list <- future_map(.x = datasets, .f = function(ds){
  # get all datasets except ds
  Input_subset.tse <- Input_parkinson_usableDatasets.tse[,
    Input_parkinson_usableDatasets.tse$study_name != ds]
  
  # create a siamcat object
  sc.obj.train <- siamcat(
    feat  = assay(Input_subset.tse, params$assay_to_test),
    meta  = as.data.frame(colData(Input_subset.tse)),
    label = 'parkinson_for_analysis',
    case  = "PD"
  )
  # normalization step is necessary, but it needs to be bypassed.
  # any normalization shoudl be done in advance
  sc.obj.train <- normalize.features(sc.obj.train, norm.method = 'pass',
                                     feature.type = "original")
  
  # create training data splits for a less biased model training
  sc.obj.train <- create.data.split(sc.obj.train,
    num.folds   = 5,
    num.resample = 5,
    inseparable = NULL
  )
  
  # train one random forest model that is an avergage of the splits above
  sc.obj.train <- train.model(sc.obj.train,
    method       = "randomForest",
    feature.type = "normalized",
    verbose      = 0,
    param.set    = list(ntree = 2000, mtry = 10)
  )
  
  return(sc.obj.train)
}) %>% 
  set_names(datasets)

t1_train_siamcat <- Sys.time()

# test left-out dataset
t0_test_siamcat <- Sys.time()
tested_models_siamcat.list <- future_imap(trained_models_siamcat.list, function(sc.obj.train, ds){

  Input_subset2.tse <- Input_parkinson_usableDatasets.tse[,
      Input_parkinson_usableDatasets.tse$study_name == ds]
    
    sc.obj.test <- siamcat(
      feat  = assay(Input_subset2.tse, params$assay_to_test),
      meta  = as.data.frame(colData(Input_subset2.tse)),
      label = 'parkinson_for_analysis',
      case  = "PD"
    )
    
    sc.obj.test <- normalize.features(sc.obj.test, norm.method = 'pass',
                                      feature.type = "original")
    
    sc.obj.test <- make.predictions(sc.obj.train, siamcat.holdout = sc.obj.test)
    sc.obj.test <- evaluate.predictions(sc.obj.test)
    
    # Extract predictions and true labels for pROC
    # pred_matrix rows = samples, cols = resamples; average across resamples
    pred_probs <- rowMeans(pred_matrix(sc.obj.test))
    
    # True binary labels: SIAMCAT  +1/-1, convert to 1/0
    true_labels_binary <- label(sc.obj.test)$label
    true_labels_binary[true_labels_binary == -1] <- 0
    
    # Build ROC with pROC and get CI
    roc_obj <- roc(response  = true_labels_binary,
                   predictor = pred_probs,
                   direction = "<",   # higher score = more likely case
                   quiet     = TRUE)
    
    # DeLong 95% CI (fast, no bootstrapping needed for CI)
    ci_obj  <- ci.auc(roc_obj, method = "delong")
    
    # p-value of PD-classification probabilities vs Control classification probabilities
    # if probabilities are high for true cases and lower for false cases, significance is higher
    # pROC's roc.test compares two ROC curves; compare against a null curve
    pd_scores  <- pred_probs[true_labels_binary == 1]
    control_scores  <- pred_probs[true_labels_binary == 0]
    wilcox_res <- wilcox.test(pd_scores, control_scores, alternative = "greater")
    
    return(data.frame(
        AUROC = as.double(auc(roc_obj)),
        AUROC_lowerCI = as.double(ci_obj[1]),   # 2.5%
        AUROC_upperCI = as.double(ci_obj[3]),   # 97.5%
        p_value_wilcox = wilcox_res$p.value
      ))}
  )
t1_test_siamcat <- Sys.time()

# create a results tibble
rf_results_siamcat.df <- bind_rows(tested_models_siamcat.list, .id = "TestDataset")
```

```{r}
# save all training models as .rds
saveRDS(tested_models_siamcat.list, file = file.path(outdir, sprintf("rf_training-PD_ctrl-LODO_%s_%s.rds", "SIAMCAT", params$assay_to_test)))
# save all polished results
write_tsv(rf_results_siamcat.df, file.path(outdir, sprintf("rf_results-PD_ctrl-LODO_%s_%s.tsv", "SIAMCAT", params$assay_to_test)))

datatable(rf_results_siamcat.df)
```

# Write clean runtimes

```{r}
runtimes.df <- data.frame(
  step = c("train", "test", "train", "test"),
  method = c("ranger", "ranger", "siamcat", "siamcat"),
  runtime_s = as.numeric(c(
    difftime(t1_train_ranger, t0_train_ranger, units = "secs"), 
    difftime(t1_test_ranger, t0_test_ranger, units = "secs"), 
    difftime(t1_train_siamcat, t0_train_siamcat, units = "secs"), 
    difftime(t1_test_siamcat, t0_test_siamcat, units = "secs"))
  ))

write_tsv(runtimes.df, file.path(outdir, sprintf("runtimes_of_both_methods_%s.tsv", params$assay_to_test)))
```

# The quick ranger implementation makes little difference in our case but is much faster

```{r}
runtimes.df <- runtimes.df %>% 
  mutate(step = factor(step, levels = c("train", "test")))

results_joined <- inner_join(rf_results_ranger.df, rf_results_siamcat.df, by = "TestDataset", suffix = c(".ranger", ".siamcat"))

datatable(results_joined)
```

## SIAMCAT does not dramatically improve accuracy, and it takes orders of magnitude longer

```{r, fig.height=6, fig.width=10}

auroc_scatter <- ggplot(results_joined, aes(x  = AUROC.ranger, y = AUROC.siamcat)) + 
  ggrepel::geom_text_repel(aes(label = TestDataset)) +
  geom_errorbarh(aes(xmin = AUROC_lowerCI.ranger, xmax = AUROC_upperCI.ranger), color = "skyblue2") +
  geom_errorbar(aes(ymin = AUROC_lowerCI.siamcat, ymax = AUROC_upperCI.siamcat), color = "gold3") +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color= "gray30", lty = "dashed") + 
  labs(
    caption = "errorbars are 95% CI of AUROC (DeLong method)"
  )

runtimes.col <- ggplot(runtimes.df, aes(x = step, y = runtime_s, fill = method)) + 
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("skyblue2", "gold3")) +
  labs(
    x = "Modeling step",
    y = "Runtime (s)"
  )

plot_summary.gg <- ggarrange(auroc_scatter, runtimes.col, widths = c(1, 0.5), common.legend = TRUE)

ggsave(plot = plot_summary.gg, filename = file.path(outdir, sprintf("ranger_vs_siamcat_%s.svg", params$assay_to_test)), height = 6, width = 10)

plot_summary.gg
```

# Session Info

```{r}
sessionInfo()
```
