---
title: "Parkinson Meta-Analysis with Random Forest LODO and sample size saturation"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_tool: "MetaPhlAn"
  assay_to_test: "relative_abundance"
  preval_thresh: 0.05
  detect_thresh: 1e-5
---

The reasoning for this analysis is trying to see if predictions by Random Forest
can improve if more samples were collected.

To do so, in each LODO iteration we will try to predict the PD-Control set using
any combination possible of the remaining datasets, including single, pairs, groups of 3, ...

We will record the number of samples in each training set and the AUC obtained 
with that test set.

```{r setup, message=FALSE, warning=FALSE}

# create base directory output
outdir <- file.path("results", "03-RF-LODO-PD-vs-Control", params$biobakery_tool, "sample-size-gradient")
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

library(knitr)
opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(mia)
library(magrittr)
library(tidyverse)
library(furrr)
# Random forest packages
library(ranger)
library(pROC)

# for table visualizations
library(DT)

# for loading tree summarized experiment
library(biobakeryUtils)

# for summary plots
library(ggpubr) 
theme_set(theme_pubr())

Input.tse <- read_TSE_from_dir(sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_complete-%s.tse", params$biobakery_tool))
```

```{r filter dataset for parkinson meta-analysis}
# Parkinson as target condition
Input_parkinson.tse <- Input.tse[, Input.tse$target_condition == "Parkinson Disease"]
Input_parkinson.tse <- Input_parkinson.tse[rowSums(assay(Input_parkinson.tse)) > 0, ]

# drop unused level of the parkinson_for_analysis
Input_parkinson.tse$parkinson_for_analysis <- droplevels(Input_parkinson.tse$parkinson_for_analysis)

# Each dataset should have at least 30 samples in cases and 30 in controls
ok_datasets <- Input_parkinson.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  filter(
    sum(parkinson_for_analysis == "Control") > 10,
    sum(parkinson_for_analysis == "PD") > 10
  )

## filter TSE to restrict to 'ok' datasets 
Input_parkinson_usableDatasets.tse <- Input_parkinson.tse[,Input_parkinson.tse$study_name %in% unique(ok_datasets$study_name)]

# ## Prevalence and detection filter
# Input_parkinson_usableDatasets.tse <- subsetByPrevalent(Input_parkinson_usableDatasets.tse, prevalence = params$preval_thresh, detection = params$detect_thresh)

Input_parkinson_usableDatasets.tse
```

```{r}
kable(table(Input_parkinson_usableDatasets.tse$study_name, Input_parkinson_usableDatasets.tse$parkinson_for_analysis, useNA = "always"))
```

# Machine Learning (RF) of PD vs controls (leave one out) with `ranger`

  0 - Data were pre-normalized with the same approach 
  as the previous meta-analysis
  
  1 - Train microbiome on all vs one dataset
  
  2 - test on the left-out dataset
  
  3 - Repeat for all

Important: `ranger` doesn't seem to have a cross-validation step in the training.
SIAMCAT has it, and it is likely what is taking so long to complete.

```{r rf classification with ranger}
plan(multisession, workers = 10)

feat_matrix <- assay(Input_parkinson_usableDatasets.tse, params$assay_to_test)
metadata    <- as.data.frame(colData(Input_parkinson_usableDatasets.tse))
datasets    <- unique(metadata$study_name)

t0_ranger <- Sys.time()

trained_models_ranger.list <- sapply(datasets, function(ds) {
    
    #training datasets
    training_datasets <- setdiff(datasets, ds)
    
    training_datasets_seq_combn <- lapply(seq_along(training_datasets), function(i) combn(training_datasets, i, simplify = FALSE)) %>% 
      unlist(recursive = FALSE)
    
    rf_results <- future_map(training_datasets_seq_combn, function(dss) {
      # prepare data for training
    train_idx      <- metadata$study_name %in% dss
    train_features <- t(feat_matrix[, train_idx])
    train_labels   <- factor(metadata$parkinson_for_analysis[train_idx])  
    
    # train rf model on all datasets except the dataset(s)
    rf_model <- ranger(
      x           = train_features,
      y           = train_labels,
      num.trees   = 2000,
      mtry        = 10,
      probability = TRUE,
      num.threads = 6,
      seed        = 1234
    )
    
    # create test sets
    test_idx      <- metadata$study_name == ds
    test_features <- t(feat_matrix[, test_idx])
  
    # true binary labels: 1 = PD, 0 = Control (mirrors SIAMCAT conversion)
    true_labels_binary <- as.integer(
      metadata$parkinson_for_analysis[test_idx] == "PD"
    )
  
    # get prediction probabilities
    pred_probs <- predict(rf_model, data = test_features)$predictions[, "PD"]
    
    # DeLong 95% Confidence intervals of AUC/AUROC
    roc_obj <- roc(response  = true_labels_binary,
                 predictor = pred_probs,
                 direction = "<",
                 quiet     = TRUE)
    
    ci_obj  <- ci.auc(roc_obj, method = "delong")
    
    return(tibble(
      trainingDatasets = paste(dss, collapse = ";"), 
      nTrainingDatasets = length(dss), 
      nTrainingSamples = sum(train_idx), 
      AUROC = as.double(auc(roc_obj)), 
      AUROC_lowerCI = as.double(ci_obj[1]), 
      AUROC_upperCI = as.double(ci_obj[3]))
      )
    }) %>% bind_rows()
    }, 
    simplify = FALSE, USE.NAMES = TRUE
    )

trained_models_ranger.df <- trained_models_ranger.list %>% 
  bind_rows(.id = "testDataset")

t1_ranger <- Sys.time()

t1_ranger - t0_ranger
```

```{r save ranger results}
# save all polished results
write_tsv(trained_models_ranger.df, file.path(outdir, sprintf("sample_size_gradient_res_%s_%s.tsv", "ranger", params$assay_to_test)))
```

```{r create plot results, warning=FALSE, message=FALSE}
sample_gradient.gg <- trained_models_ranger.df %>% 
  ggplot(aes(x = nTrainingSamples, y = AUROC, color = factor(nTrainingDatasets), key = trainingDatasets)) + 
  geom_point() + 
  facet_wrap(~testDataset) + 
  scale_color_viridis_d(direction = -1)

sample_gradient.gg
```

```{r}
plotly::ggplotly(sample_gradient.gg)
```


```{r save plot results}
# svg
ggsave(plot = sample_gradient.gg, filename = file.path(outdir, sprintf("sample_size_gradient_%s_%s.svg", "ranger", params$assay_to_test)), width = 9, height = 9)

# png
ggsave(plot = sample_gradient.gg, filename = file.path(outdir, sprintf("sample_size_gradient_%s_%s.png", "ranger", params$assay_to_test)), width = 9, height = 9)
```

# Session Info

```{r session info}
sessionInfo()
```
