---
title: "Constipated microbiome in healthy (non-PD) individuals"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_tool: "HUMAnN"
  preval_thresh: 0.05
  detect_thresh: 1e-5
---

```{r setup, message=FALSE, warning=FALSE}
# ------ prepare outdir --------
outdir <- file.path("results", "04-Constipation-healthy-metaAnalysis", params$biobakery_tool)
dir.create(outdir, recursive = TRUE)
# ------------------------------

library(meta)
library(limma)
library(biobakeryUtils)
library(mia)
library(openxlsx2)
library(tidyverse)
library(knitr)

Input.tse <- read_TSE_from_dir(sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_complete-%s.tse", params$biobakery_tool))

Input.tse
```

## Prepare dataset for healthy constipated analysis

```{r filter dataset for constipation meta-analysis}
# filter for disease == Healthy
Input_constipation_healthy.tse <- Input.tse[, Input.tse$disease == "Healthy"]

# filter for available constipation information
Input_constipation_healthy.tse <- Input_constipation_healthy.tse[,!is.na(Input_constipation_healthy.tse$constipation_for_analysis)]

# Prevalence and detection filter
Input_constipation_healthy.tse <- subsetByPrevalent(Input_constipation_healthy.tse, prevalence = params$preval_thresh, detection = params$detect_thresh)

Input_constipation_healthy.tse
```

```{r samples by study_name}
kable(table(Input_constipation_healthy.tse$study_name, Input_constipation_healthy.tse$constipation_for_analysis, useNA = "always"))
```

# Constipation fixed effects meta-analysis

## Single dataset analysis with `limma-voom`

```{r limma regression}
per_dataset_DA_results_limma.tb <- Input_constipation_healthy.tse %>% 
  splitOn(group = "study_name") %>% 
  lapply(function(x.tse) {
    x.tse <- x.tse[rowSums(assay(x.tse)) > 0,]
    return(limmaTSE(tse = x.tse, assay.type = "counts", voom.transform = TRUE, formula = ~ constipation_for_analysis, coef = "constipation_for_analysisSlow", p.adj_method = "BH"))
  }
  ) %>% bind_rows(.id = "study_name")
```

## Single dataset analysis with `Maaslin2`

```{r maaslin2 regression, include=FALSE}
per_dataset_DA_results_maaslin2.tb <- Input_constipation_healthy.tse %>%
  splitOn(group = "study_name") %>%
  lapply(function(x.tse) {
    x.tse <- x.tse[rowSums(assay(x.tse, "relative_abundance")) > 0, ]
    return(
      maaslin2TSE(
        tse = x.tse,
        assay.type = "relative_abundance",
        outdir = tempfile(pattern = "maaslin2_"),
        min_variance = 0,
        transform = "LOG",
        max_significance = 0.1,
        fixed_effects = "constipation_for_analysis",
        correction = "BH",
        outdir_delete = TRUE,
        normalization = "NONE",
        min_prevalence = 0,
        min_abundance = 0,
        plot_heatmap = FALSE,
        plot_scatter = FALSE,
        save_scatter = FALSE,
        standardize = FALSE
      )
    )
  }) %>% bind_rows(.id = "study_name") %>% 
  filter(ExposureValue == "constipation_for_analysisSlow") %>% 
  # readjust the p-value because we are interested only in one comparison
  group_by(study_name) %>% 
  mutate(P.Value.adj = p.adjust(P.Value, "BH")) %>% 
  ungroup()
```

## Fixed effects Meta-analysis 

```{r, fixed effects meta-analysis limma}
MetaAnalysis_fixedEff_results_limma <- metaAnalyze(res.df = per_dataset_DA_results_limma.tb, grouping.col = "study_name", beta.col = "Beta", se.col = "SE", method = "FE", test = "knha")
```

```{r, fixed effects meta-analysis maaslin2}
MetaAnalysis_fixedEff_results_maaslin2 <- metaAnalyze(res.df = per_dataset_DA_results_maaslin2.tb, grouping.col = "study_name", beta.col = "Beta", se.col = "SE", method = "FE", test = "knha")
```

## Random effects Meta-analysis 

```{r, random effects meta-analysis limma}
MetaAnalysis_randEff_results_limma <- metaAnalyze(res.df = per_dataset_DA_results_limma.tb, grouping.col = "study_name", beta.col = "Beta", se.col = "SE", method = "REML", test = "knha")
```

```{r, random effects meta-analysis maaslin2}
MetaAnalysis_randEff_results_maaslin2 <- metaAnalyze(res.df = per_dataset_DA_results_maaslin2.tb, grouping.col = "study_name", beta.col = "Beta", se.col = "SE", method = "REML", test = "knha")
```

# Write all tables

```{r initialize file index for writing}
outputID <- 1
```

## limma results 

```{r}
outfile <- file.path(outdir, sprintf("%02d-%s_per_dataset_DA_results_limma.tsv", outputID, params$biobakery_tool))
write_tsv(per_dataset_DA_results_limma.tb, file = outfile)
outputID <- outputID + 1
```

## Maaslin2 results

```{r}
outfile <- file.path(outdir, sprintf("%02d-%s_per_dataset_DA_results_maaslin2.tsv", outputID, params$biobakery_tool))
write_tsv(per_dataset_DA_results_maaslin2.tb, file = outfile)
outputID <- outputID + 1
```

## Fixed effects meta-analysis on limma output

```{r, write constipation limma fixed meta-analysis}
outfile <- file.path(outdir, sprintf("%02d-%s_MetaAnalysis_fixedEff_results_limma.xlsx", outputID, params$biobakery_tool))
openxlsx2::write_xlsx(MetaAnalysis_fixedEff_results_limma, file = outfile)
outputID <- outputID + 1
```

## Fixed effects meta-analysis on maaslin2 output

```{r, write constipation maaslin2 fixed meta-analysis}
outfile <- file.path(outdir, sprintf("%02d-%s_MetaAnalysis_fixedEff_results_maaslin2.xlsx", outputID, params$biobakery_tool))
openxlsx2::write_xlsx(MetaAnalysis_fixedEff_results_maaslin2, file = outfile)
outputID <- outputID + 1
```

## Random effects Meta-analysis on limma output

```{r, write constipation limma random meta-analysis}
outfile <- file.path(outdir, sprintf("%02d-%s_MetaAnalysis_randEff_results_limma.xlsx", outputID, params$biobakery_tool))
openxlsx2::write_xlsx(MetaAnalysis_randEff_results_limma, file = outfile)
outputID <- outputID + 1
```

## Random effects Meta-analysis on maaslin2 output

```{r, write constipation maaslin2 random meta-analysis}
outfile <- file.path(outdir, sprintf("%02d-%s_MetaAnalysis_randEff_results_maaslin2.xlsx", outputID, params$biobakery_tool))
openxlsx2::write_xlsx(MetaAnalysis_randEff_results_maaslin2, file = outfile)
outputID <- outputID + 1
```

# Session info

```{r session info}
sessionInfo()
```
