---
title: "Healthy constipation signature and score"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  biobakery_tool: "HUMAnN"
  preval_thresh: 0.05
  detect_thresh: 1e-5
---

```{r setup, message=FALSE, warning=FALSE}
# ------ prepare outdir --------
outdir <- file.path("results", "05-Constipation-score", params$biobakery_tool)
dir.create(outdir, recursive = TRUE)
# ------------------------------

library(biobakeryUtils)
library(SummarizedExperiment)
library(tidyverse)
library(knitr)
library(ggpubr)
library(Maaslin2)
library(furrr)

# infer the tested unit from the biobakery tool
tested_unit <- ifelse(params$biobakery_tool == "MetaPhlAn", "SGB", "pathway")

Input.tse <- read_TSE_from_dir(sprintf("~/parkinsonMetaAnalysis_data/03-analysis_ready_datasets/01-dataset_complete-%s.tse", params$biobakery_tool))

set.seed(1234)

Input.tse
```

## Calculate and save two different constipation scores

```{r prepare constipation score vectors}
# load constipation DA results
bluePoo_constipation_results <- read_tsv(file.path("results", "04-Constipation-healthy-metaAnalysis", sprintf("%s/02-%s_per_dataset_DA_results_maaslin2.tsv", params$biobakery_tool, params$biobakery_tool))) %>% 
  filter(study_name == "AsnicarF_2021")

bluePoo_score <- bluePoo_constipation_results %>% 
  split.data.frame(sign(.$t_stat)) %>% 
  lapply(function(x) {
    t_vec <- x$t_stat
    names(t_vec) <- x[[tested_unit]]
    return(t_vec)
  }
   ) %>% Reduce(c, .)

bluePoo_score_only_signif <- bluePoo_constipation_results %>% 
  filter(P.Value.adj < 0.1) %>% 
  split.data.frame(sign(.$t_stat)) %>% 
  lapply(function(x) {
    t_vec <- x$t_stat
    names(t_vec) <- x[[tested_unit]]
    return(t_vec)
  }
   ) %>% Reduce(c, .)
```

```{r calculate constipation score}

assay_table.mtx <- assay(Input.tse, "relative_abundance")

# calculate constipation score 1 - sum of relative abundance of only significant features

## create an intermediate score that comprises only singnificant associations from the discovery dataset
score_intermediate <- bluePoo_score_only_signif[match(rownames(assay_table.mtx), names(bluePoo_score_only_signif))]
names(score_intermediate) <- rownames(assay_table.mtx)
## replace NAs introduced by match with 0 because that feature is not part of 
## the score set of bacteria
score_intermediate[is.na(score_intermediate)] <- 0

constipation_score1 <- apply(assay_table.mtx, 2, function(x) sum(x * sign(score_intermediate)))
#-----------------------------------------------------

# calculate constipation score 2

## create an intermediate score that comprises ALL features found in target dataset
score_intermediate <- bluePoo_score[match(rownames(assay_table.mtx), names(bluePoo_score))]
names(score_intermediate) <- rownames(assay_table.mtx)
## replace NAs introduced by match with 0 because that feature is not part of 
## the score set of bacteria
score_intermediate[is.na(score_intermediate)] <- 0

constipation_score2 <- apply(assay_table.mtx, 2, function(x) sum(x * score_intermediate))
#-----------------------------------------------------

# append scores to wallen metadata
Input.tse$constipation_score1 <- constipation_score1
Input.tse$constipation_score2 <- constipation_score2
#-----------------------------------------------------
```

## Save scores as supplements

```{r}
constipation_scores <- Input.tse %>% 
  colData() %>% 
  as.data.frame() %>% 
  dplyr::select(uuid, constipation_score1, constipation_score2)

write_tsv(constipation_scores, sprintf("../05-Supplements/Supplementary Tables/constipation_score_%s.tsv", params$biobakery_tool))
```

## Mediation of Constipation on the PD effects on the Microbiome (on WallenZD_2021)

Dataset: `WallenZ_2022`
Outcome: SGBs (log-normal continuous)
Exposure: PD (binary)
Mediator: constipation score (binary)

In the `mediation::mediate` function, non-parametric resampling would be more 
accurate (`boot = TRUE`) than quasi-Bayesian Monte Carlo (`boot = FALSE`).

The latter, with the setup as it is now, returns an error:

```
Error in if (xhat == 0) out <- 1 else { : 
  missing value where TRUE/FALSE needed
```

likely a zero-variance error. Maybe be more restrictive with variance filtering 
during the maaslin2 call. As proof of concept, I am leaving `boot = FALSE`. We
can refine later

Classical mediation framework: https://pmc.ncbi.nlm.nih.gov/articles/PMC6180946/
example usage with continuous outcome but binary mediator and exposure:
https://imai.fas.harvard.edu/projects/files/mediationReview.pdf

Below the classical 3 models that are used to estimate mediation

$Exposure~on~outcome: Microbiome \sim Parkinson$

$Exposure~on~mediator: Constipation \sim Parkinson$

$Exposure~and~mediator~on~outcome: Microbiome \sim Parkinson + Constipation$

### Maaslin2 set of regressions

```{r mediation regressions}

WallenZD_2022.tse <- Input.tse[, Input.tse$study_name == "WallenZD_2022"]
WallenZD_constip_2022.tse <- WallenZD_2022.tse[, !is.na(WallenZD_2022.tse$constipation_for_analysis)]
WallenZD_constip_2022.tse <- WallenZD_constip_2022.tse[rowSums(assay(WallenZD_constip_2022.tse)) > 0,]

t0_mediation <- Sys.time()

set.seed(1234)
outdir_mediation <- file.path(outdir, "mediation_Wallen")
dir.create(outdir_mediation, recursive = TRUE)

wallen_metadata.df <- as.data.frame(colData(WallenZD_constip_2022.tse))
wallen_assay <- as.data.frame(t(assay(WallenZD_constip_2022.tse, "relative_abundance")))

# Mediator Model (M)
m_model <- glm(constipation_for_analysis ~ parkinson_for_analysis, binomial(link = "logit"), data = wallen_metadata.df)

# Full model (F)
output_captured <- capture.output(
  Maaslin2(
    input_data = wallen_assay,
    input_metadata = wallen_metadata.df,
    output = outdir_mediation,
    save_models = TRUE,
    plot_heatmap = FALSE,
    plot_scatter = FALSE,
    min_abundance = 0,
    min_prevalence = 0.05,
    min_variance = 0,
    normalization = "none",
    transform = "LOG",
    analysis_method = "LM",
    fixed_effects = c("parkinson_for_analysis", "constipation_for_analysis")
  )
)
```

### Mediation analysis

```{r mediation}
data_for_regressions <- read_tsv(file.path(outdir_mediation, "features", "filtered_data_norm_transformed.tsv")) %>% 
  dplyr::rename(uuid = feature)

fit_data <- dplyr::left_join(wallen_metadata.df, data_for_regressions, by = "uuid")

features <- colnames(data_for_regressions[,-1])
# --- Parallelization setup ---
plan(multisession, workers = 8)

mediations.list <- future_map(
  features,
  function(tx) {
    
    y_fit <- lm(
      as.formula(sprintf("%s ~ parkinson_for_analysis + constipation_for_analysis", tx)),
      data = fit_data
    )
    
    set.seed(1234)
    
    mediation <- suppressWarnings(
      mediation::mediate(
        model.m  = m_model,
        model.y  = y_fit,
        treat    = "parkinson_for_analysis",
        mediator = "constipation_for_analysis",
        boot     = FALSE,
        sims     = 1000
      )
    )
    
    return(mediation)
  },
  # crucial step, import only essential things into the R sessions
  # this remains a little inefficient because all models are exported in each 
  # session, taking 500 MB in each 
  .options = furrr_options(
    seed = TRUE,
    globals = list(
      fit_data = fit_data,
      m_model = m_model
    ),
    packages = c("mediation")
  )
)

names(mediations.list) <- features

# stop parallel backend
plan(sequential)

t1_mediation <- Sys.time()
rm(output_captured)
```

```{r}
t1_mediation - t0_mediation
```

```{r}
results_clean <- bind_rows(lapply(mediations.list, tidy_mediation), .id = tested_unit)
# mutate(Effect = factor(Effect))
# levels(results_clean$Effect) <- levels(results_clean$Effect)[c(4,2,1,3)]
# write full results
write_tsv(results_clean, file = file.path(outdir_mediation, "mediation_analysis_results_full.tsv"))
```

# Session info

```{r session info}
sessionInfo()
```
