---
title: "Study tables, MetaPhlAn alpha and beta diversities"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  pMD_data_type: "relative_abundance"
  pseudocount: 1e-8
  transform: "asin_sqrt"
  preval_thresh: 0.05
  detect_thresh: 1e-5
---

# Data and environment preparation 

```{r setup, message=FALSE, warning=FALSE}
source("../00-PD-Control-meta-analysis-MetaPhlAn-data-prep.R", chdir = TRUE)
```

# Table 1 - dataset x PD, unavailable covariates limit meta-analysis strength

## All samples available after filtering

```{r table of samples per condition}
library(gtsummary)

table1_complete <- colData(Input.tse) %>% 
  as.data.frame() %>% 
  select(`Study` = study_name, Country, `Age Gr.` = age_group, Sex = sex, `N. diseases` = number_diseases, Parkinson_1_0) %>% 
   tbl_summary(
    by = Parkinson_1_0, # Optional: stratify by a variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing_text = "Missing"
  ) %>%
  add_overall() %>% # Add an overall column
  modify_header(label ~ "**Characteristic**") # Customize header


```

## Datasets with available age and sex information

```{r table of samples per condition with both sex and age_group available}
table1_availSex <- colData(Input.tse) %>% 
  as.data.frame() %>% 
  filter(complete.cases(select(., age_group, sex))) %>% 
  select(`Study` = study_name, Country, `Age Gr.` = age_group, Sex = sex, `N. diseases` = number_diseases, Parkinson_1_0) %>% 
   tbl_summary(
    by = Parkinson_1_0, # Optional: stratify by a variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing_text = "Missing"
  ) %>%
  add_overall() %>% # Add an overall column
  modify_header(label ~ "**Characteristic**") # Customize header
```

## Datasets with at least 30 individuals in PD and Control groups

```{r table of samples per condition with sufficient case/ctrl samples}
ok_datasets <- Input.tse@colData %>% 
  as.data.frame() %>% 
  group_by(study_name) %>% 
  filter(
    sum(Parkinson_1_0 == 0) > 30,
    sum(Parkinson_1_0 == 1) > 30
  )

Input_usableDatasets.tse <- Input.tse[,Input.tse@colData$study_name %in% unique(ok_datasets$study_name)]

# filter all-o rows again
Input_usableDatasets.tse <- Input_usableDatasets.tse[rowSums(assay(Input_usableDatasets.tse)) > 0,]

table1_ok_Parkinson <- colData(Input_usableDatasets.tse) %>% 
  as.data.frame() %>% 
  filter(complete.cases(select(., age_group, sex))) %>% 
  select(`Study` = study_name, Country, `Age Gr.` = age_group, Sex = sex, `N. diseases` = number_diseases, Parkinson_1_0) %>% 
   tbl_summary(
    by = Parkinson_1_0, # Optional: stratify by a variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing_text = "Missing"
  ) %>%
  add_overall() %>% # Add an overall column
  modify_header(label ~ "**Characteristic**") # Customize header

Input_usableDatasets.tse
```

# Alpha diversities

```{r calculate alpha diverity metrics and make linear models}
alphas <- c("observed_richness", "shannon_diversity", "faith")
Input_usableDatasets.tse <- AddPhyloTree_to_mpa_tse(Input_usableDatasets.tse, chocophlan_datestamp = "202403")

# Calculate alphas
Input_usableDatasets.tse <- addAlpha(Input_usableDatasets.tse, assay.type = "relative_abundance", index = alphas)
```

## PCoAs

PCoAs depend on the input data transformation, this is an attempt to incorporate
Multiple options

```{r curate beta params}
if(params$transform == "asin_sqrt"){
  assay(Input_usableDatasets.tse, "asin_sqrt") <- asin(sqrt(assay(Input_usableDatasets.tse)))
  beta.method = "euclidean"
}

if(params$transform == "clr"){
  Input_usableDatasets.tse <- transformAssay(Input_usableDatasets.tse, assay.type = "relative_abundance", method = params$transfor, pseudocount = params$pseudocount)
  beta.method = "euclidean"
}

if(params$transform == "relative_abundance" | is.null(params$transform)){
  beta.method = "bray"
}

beta.name <- paste(params$transform, beta.method, sep = "_")
```

```{r MDS plots}
Input_usableDatasets.tse <- addMDS(Input_usableDatasets.tse, assay.type = params$transform, method = beta.method, name = beta.name)

variances_explained <- attr(reducedDim(Input_usableDatasets.tse, beta.name), "eig")
prop_variances_explained <- variances_explained/sum(variances_explained)

betadiv_data.df <- cbind.data.frame(as.data.frame(colData(Input_usableDatasets.tse)), reducedDim(Input_usableDatasets.tse, beta.name) %>% as.data.frame() %>% set_names(paste("MDS", 1:ncol(.))))

colorVars <- c("Country", "study_name", "Parkinson_1_0")

PCoA_plots.list <- map(colorVars, function(colorVar) {
  betadiv_data.df %>%
    ggplot(aes(x = `MDS 1`, y = `MDS 2`, color = !!sym(colorVar), fill = !!sym(colorVar))) + 
  geom_point(size = 2.5, stroke = 1, color = "black") +
  geom_point(size = 2.5) +
  ggConvexHull::geom_convexhull(show.legend = FALSE, alpha = 0.2) + 
  ggsci::scale_color_jco() +
  ggsci::scale_fill_jco() +
  theme(legend.position = "top", legend.text = element_markdown()) +
  labs(color = colorVar,
       fill = colorVar,
       x = paste0("PCoA Axis 1 [", round(prop_variances_explained[1]*100, 1), "%]"),
       y = paste0("PCoA Axis 2 [", round(prop_variances_explained[2]*100, 2), "%]")
  )
}) %>% 
  set_names(sprintf("MDS_%s_color_%s", beta.name, colorVars))

PCoA_plots.list[[3]]
```

```{r PERMANOVA, cache=TRUE}
PERMANOVA_name <- paste(beta.name, "PERMANOVA", sep = "_")

Input_usableDatasets.tse <- addPERMANOVA(
  Input_usableDatasets.tse,
  assay.type = params$transform,
  method = beta.method,
  name = PERMANOVA_name,
  formula = x ~ study_name + Parkinson_1_0 + reads_processed,
  by = "margin",
  permutations = 2000, 
  parallel = 8
)

PERMANOVA_to_table(Input_usableDatasets.tse, name = PERMANOVA_name)
```

# Output Figures (as `.rds`)

```{r figures outdir setup}
outdir_figures <- file.path(outdir, "Figures")
dir.create(outdir_figures)
```

## List of PCoA plots colored by different colData columns

```{r save MDS plots as .rds}
imap(PCoA_plots.list, function(plot.gg, pltName){
  saveRDS(plot.gg, file = sprintf("%s/01-%s.rds", outdir_figures, pltName))
})
```

# Output Tables (multiple formats)

```{r tables outdir setup}
outdir_tables <- file.path(outdir, "Tables")
dir.create(outdir_tables)
```

## Filtering steps - samples and features excluded (as `.tsv`)

```{r steps of samples and features exclusions}
write_tsv(rbind(filtering_steps.tb, c("Enough Cases and Controls per dataset", dim(Input_usableDatasets.tse))), file = sprintf("%s/01-filtering-samples-and-features-excluded.tsv", outdir_tables))
```

## Three versions of study tables - overview of datasets and covariates (as `.docx`)

```{r write demographic tables}
as_flex_table(table1_complete) %>% flextable::save_as_docx(path = sprintf("%s/02-table1-complete.docx", outdir_tables))

as_flex_table(table1_availSex) %>% flextable::save_as_docx(path = sprintf("%s/02-table1-availSex.docx", outdir_tables))

as_flex_table(table1_ok_Parkinson) %>% flextable::save_as_docx(path = sprintf("%s/02-table1-okParkinson.docx", outdir_tables))
```

## Alpha diversity tables for future figures and modeling

```{r write alpha diversity data for future plots and stats}
data_for_alpha_calculations <- colData(Input_usableDatasets.tse) %>% 
  as.data.frame() %>% 
  select(all_of(alphas), Parkinson_1_0, age_group, number_diseases, reads_processed, study_name) %>% 
write_tsv(file = sprintf("%s/03-alpha_diversity_data.tsv", outdir_tables))
```

## Alpha diversity modeling results

```{r lm of alpha vs PD and other covariates}
# Model alpha diversity including covariates
alpha_regression_pd_study.lm.list <- sapply(alphas, function(x) {
  lm(as.formula(sprintf("%s ~ Parkinson_1_0 + study_name + number_diseases", x)), data = data_for_alpha_calculations)
}, simplify = FALSE) %>% 
  lapply(broom::tidy) %>%
  lapply(mutate, FDR = p.adjust(p.value, "BH"))

bind_rows(alpha_regression_pd_study.lm.list, .id = "alpha_metric") %>% 
  write_tsv(file = sprintf("%s/04-alpha_diversity_models_joined.tsv", outdir_tables))
```

## PERMANOVA table

```{r PERMANOVA calculation based on beta-diversity}
biobakeryUtils::PERMANOVA_to_table(Input_usableDatasets.tse, name = PERMANOVA_name) %>% 
  write_tsv(file = sprintf("%s/05-%s.tsv", outdir_tables, PERMANOVA_name))
```

# Session Info

```{r}
sessionInfo()
```
