---
title: "Constipated microbiome in healthy (non-PD) individuals"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
params:
  pMD_data_type: "relative_abundance"
  pseudocount: 1e-8
  transform: "asin_sqrt"
  prevalThresh: 0.05
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(mia)
library(magrittr)
library(tidyverse)
library(parkinsonsMetagenomicData)
library(ggtext)

library(ggpubr)
library(meta)

# for table visualizations
library(DT)
library(kableExtra)
library(openxlsx2)

library(limma)
library(biobakeryUtils)
theme_set(theme_bw())

source("../R/lm_standardization_utility_functions.R")
```

# `AsnicarF_2021`

## Load private metadata and define transit time groups

```{r}
# This generates a merged version to map to predict metadata
source("../Data/bluePoo/bluePoo_metadata/bluePoo_harmonize_metadata.R")

studyNames <- c("WallenZD_2022", "AsnicarF_2021")
bluePoo_vars.chr <- c("bowel_movements_last_7_days", "bristol_stool_score_average_last_3_months", "transit time (days)")
# load separate asnicar metadata
bluePoo_meta_raw.df <- read_tsv("../Data/bluePoo/bluePoo_metadata/bluePoo_metadata_ga.tsv") 

# process what we need from this metadata
AsnicarF_meta_select.df <-  bluePoo_meta_raw.df %>% 
  transmute(
    subject_id = sample_alias,
    `transit time (hours)` = `transit time (days)` * 24,
    transitTimeCut = cut(`transit time (hours)`, breaks = c(0, 14, 38, 58,max(`transit time (hours)` + 1e-6, na.rm = TRUE)), right = FALSE),
    transitTimeCutClasses = cut(`transit time (hours)`, breaks = c(0, 14, 38, 58,max(`transit time (hours)` + 1e-6, na.rm = TRUE)), right = FALSE, labels = paste0("C", 1:4)),
    transitTimeCutClassesDef =  case_when(
     transitTimeCutClasses == "C1" ~ "Fast",
     transitTimeCutClasses %in% c("C2","C3") ~ "Normal",
     transitTimeCutClasses == "C4" ~ "Slow"
     ) %>% factor(c("Normal", "Fast", "Slow"))
    )
AsnicarF_pMD.df <- read_tsv("../Data/bluePoo/bluePoo_metadata/AsnicarF_2021.tsv")

Asnicar_complete.df <- inner_join(AsnicarF_meta_select.df, AsnicarF_pMD.df, by = "subject_id")
```

```{r}
rects <- data.frame(
  xmin = c(0, 14, 38, 58),
  xmax = c(14, 38, 58, max(Asnicar_complete.df$`transit time (hours)`, na.rm = TRUE)),
  ymin = 0,
  ymax =  max(density(Asnicar_complete.df$`transit time (hours)`, na.rm = TRUE)[["y"]]),
  band = factor(c("C1", "C2", "C3", "C4"))
)

ggplot(Asnicar_complete.df, aes(x = `transit time (hours)`)) +
  geom_rect(data = rects,
            aes(xmin = xmin, xmax = xmax,
                ymin = ymin, ymax = ymax, fill = band),
            inherit.aes = FALSE,
            alpha = 0.8) +
  geom_density(color = "black", size = 0.6) +
  scale_fill_brewer(palette = "Blues")

```

```{r}
kbl(table(Asnicar_complete.df$transitTimeCutClassesDef, Asnicar_complete.df$transitTimeCutClasses,useNA = "ifany")) %>% 
  kable_styling() %>% 
  column_spec(1, bold = TRUE)
```

## Load AsnicarF_2021 pMD microbiome data (publicly available)

```{r}
# get pMD microbiome profiles 
basic.tse <- returnSamples(Asnicar_complete.df, data_type = params$pMD_data_type)

Asnicar_complete.tse <- basic.tse %>% 
  biobakeryUtils::pMD_enhance(data_type = params$pMD_data_type)%>% 
  replace_colData(Asnicar_complete.df %>% as.data.frame() %>% set_rownames(.$uuid))

# Remove UNCLASSIFIED (MetaPhlAn) |UNMAPPED (HUMAnN) |UNINTEGRATED (HUMAnN) rows
Asnicar_complete.tse <- Asnicar_complete.tse[!grepl("UN", rownames(Asnicar_complete.tse)),]

assay(Asnicar_complete.tse, "asin_hellinger") <- asin(sqrt(assay(Asnicar_complete.tse, "relabundance")))

# get only available covariate variables
Asnicar_complete_no_NA_Constipation.tse <- Asnicar_complete.tse[, !is.na(Asnicar_complete.tse@colData$transitTimeCutClassesDef)]

# remove all-0 rows after filtering
Asnicar_complete_no_NA_Constipation.tse <- Asnicar_complete_no_NA_Constipation.tse[rowSums(assay(Asnicar_complete_no_NA_Constipation.tse, "relabundance")) > 0,]

Asnicar_complete_no_NA_Constipation.tse
```

## Fit Asnicar Data (`asin(Hellinger)`)

```{r}

slow_vs_normal <- limmaFit_TSE(Asnicar_complete_no_NA_Constipation.tse, assay.type = "asin_hellinger", formula = ~ transitTimeCutClassesDef, coef = "transitTimeCutClassesDefSlow") %>% 
  arrange(P.Value.adj.BH) %>% 
  mutate(
    CohenD = cohen_d_from_lm(
    n1 = sum(Asnicar_complete_no_NA_Constipation.tse@colData$transitTimeCutClassesDef == "Slow"), 
    n2 = sum(Asnicar_complete_no_NA_Constipation.tse@colData$transitTimeCutClassesDef == "Normal"), 
    t = t_stat),
    CohenSE = cohen_d_StdErr(
      n1 = sum(Asnicar_complete_no_NA_Constipation.tse@colData$transitTimeCutClassesDef == "Slow"), 
    n2 = sum(Asnicar_complete_no_NA_Constipation.tse@colData$transitTimeCutClassesDef == "Normal"),
    d = CohenD
    )
    )

slow_vs_normal_prev5pct <- slow_vs_normal %>% 
  filter(N_not_zero/N_zero >= 0.05)

fast_vs_normal <- limmaFit_TSE(Asnicar_complete_no_NA_Constipation.tse, assay.type = "asin_hellinger", formula = ~ transitTimeCutClassesDef, coef = "transitTimeCutClassesDefFast") %>% 
  arrange(P.Value.adj.BH) %>% 
  mutate(
    CohenD = cohen_d_from_lm(
    n1 = sum(Asnicar_complete_no_NA_Constipation.tse@colData$transitTimeCutClassesDef == "Fast"), 
    n2 = sum(Asnicar_complete_no_NA_Constipation.tse@colData$transitTimeCutClassesDef == "Normal"), 
    t = t_stat),
    CohenSE = cohen_d_StdErr(
      n1 = sum(Asnicar_complete_no_NA_Constipation.tse@colData$transitTimeCutClassesDef == "Fast"), 
    n2 = sum(Asnicar_complete_no_NA_Constipation.tse@colData$transitTimeCutClassesDef == "Normal"),
    d = CohenD
    )
    )

fast_vs_normal_prev5pct <- fast_vs_normal%>% 
  filter(N_not_zero/N_zero >= 0.05)

inner_join(fast_vs_normal_prev5pct, slow_vs_normal_prev5pct, by = "FeatureID") %>% 
  ggplot(aes(x = Beta.x, y = Beta.y)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1, color = "red") + 
  geom_abline(intercept = 0, slope = -1, color = "red") + 
  stat_cor(method = "pearson") + 
  geom_smooth(method = "lm", formula = y ~ x) + 
  labs(
    title = "Effect Size scatterplot",
    subtitle = "baseline: Normal transit time", 
    x = "Fast Transit",
    y = "Slow Transit"
  )
```

Associations of fast and slow transit against normal transit return negatively correlated associations, meaning a considerable portion of bugs that are associated in one direction in the slow transit are in the opposite direction in the fast transit and vice versa.

# WallenZD_2022

```{r}
WallenZD_2022.tse <- sampleMetadata %>% 
  filter(study_name == "WallenZD_2022") %>% 
  returnSamples(data_type = params$pMD_data_type) %>% 
  pMD_enhance(data_type = params$pMD_data_type)

# remove swab samples
WallenZD_2022_noSwabs.tse <- WallenZD_2022.tse[, WallenZD_2022.tse@colData$uncurated_collection_method != "swab"]

# prune resulting table
WallenZD_2022_noSwabs.tse <- WallenZD_2022_noSwabs.tse[rowSums(assay(WallenZD_2022_noSwabs.tse)) > 0,]

# calculate asin(sqrt(relabundance))
assay(WallenZD_2022_noSwabs.tse, "asin_hellinger") <- asin(sqrt(assay(WallenZD_2022_noSwabs.tse, "relabundance")))

table(WallenZD_2022_noSwabs.tse@colData$uncurated_constipation, WallenZD_2022_noSwabs.tse@colData$control, useNA = "ifany")
```

## Fit Healthy individuals in Wallen data (`asin(Hellinger)`)

```{r}
WallenZD_2022_noSwabs_noPD.tse <- WallenZD_2022_noSwabs.tse[, WallenZD_2022_noSwabs.tse@colData$control == 'Study Control' &  WallenZD_2022_noSwabs.tse@colData$uncurated_constipation != ""]

WallenZD_2022_noSwabs_noPD.tse <- WallenZD_2022_noSwabs_noPD.tse[rowSums(assay(WallenZD_2022_noSwabs_noPD.tse)) > 0, ]

# recode constipation
WallenZD_2022_noSwabs_noPD.tse@colData$uncurated_constipation <- factor(WallenZD_2022_noSwabs_noPD.tse@colData$uncurated_constipation, levels = c("N", "Y"))

wallen_fit_Controls_constipated_vs_not_constipated <- limmaFit_TSE(WallenZD_2022_noSwabs_noPD.tse, assay.type = "asin_hellinger", formula = ~ uncurated_constipation, coef = "uncurated_constipationY") %>% 
  arrange(P.Value.adj.BH) %>% 
  mutate(CohenD = cohen_d_from_lm(
    n1 = sum(WallenZD_2022_noSwabs_noPD.tse@colData$uncurated_constipation == "Y"), 
    n2 = sum(WallenZD_2022_noSwabs_noPD.tse@colData$uncurated_constipation == "N"), 
    t = t_stat),
    CohenSE = cohen_d_StdErr(
      n1 = sum(WallenZD_2022_noSwabs_noPD.tse@colData$uncurated_constipation == "Y"),
      n2 = sum(WallenZD_2022_noSwabs_noPD.tse@colData$uncurated_constipation == "N"),
      d = CohenD
    )
    )

wallen_fit_Controls_constipated_vs_not_constipated_5perc <- wallen_fit_Controls_constipated_vs_not_constipated %>% 
   filter(N_not_zero/N_zero >= 0.05)
```

# Write tables

```{r}
outdir <- sprintf("../results/04-Constipation__%s", params$pMD_data_type)
dir.create(outdir)

write_xlsx(wallen_fit_Controls_constipated_vs_not_constipated_5perc, file = sprintf("%s/limma_regression_WallenZD_2022_ConstipatedControls.xlsx", outdir))

write_xlsx(
  list("C4 vs C2C3" = slow_vs_normal_prev5pct, "C1 vs C2C3" = fast_vs_normal_prev5pct),
  file = sprintf("%s/limma_regression_BluePoo.xlsx", outdir)
)
```

# Compare effect sizes of the two associations

```{r}
merging_cols <- colnames(wallen_fit_Controls_constipated_vs_not_constipated_5perc)[1:(which(colnames(wallen_fit_Controls_constipated_vs_not_constipated_5perc) == "N") - 1)]
constipation_joined_analyses <- inner_join(wallen_fit_Controls_constipated_vs_not_constipated_5perc, slow_vs_normal_prev5pct, by = merging_cols, suffix = c(".Wallen", ".Asnicar"))
```

## Top 10 largest deviations

```{r}
constipation_joined_analyses_sub <- constipation_joined_analyses %>% 
  mutate(absDiff = abs(Beta.Wallen - Beta.Asnicar)) %>% 
  arrange(desc(absDiff)) %>% 
  select(FeatureID, contains(c("CohenD", "CohenSE", "Beta", "SE", "adj.P.Value")), absDiffBetas = absDiff)

constipation_joined_analyses_sub %>% 
  mutate_if(is.numeric, format, scientif = TRUE, digits = 1) %>% 
  datatable()
```

# Overlap with PD meta-analysis

```{r}
meta_analysis <- read_xlsx(sprintf("../results/03_meta_analysis__%s/02-meta_analysis_%s_PD_only.xlsx", params$pMD_data_type, params$transform), sheet = "Meta-Analysis")
colnames(meta_analysis) <- gsub("RE_", "", colnames(meta_analysis))

meta_analysis_subset <- meta_analysis %>% 
  transmute(
    FeatureID, 
    Eff.Size,
    PD_assoc_metaAnalysis = case_when(
      Eff.Size > 0 & FDR < 0.1 ~ "Increased", 
      Eff.Size < 0 & FDR < 0.1 ~ "Decreased",
      FDR >= 0.1 ~ "NS",
      het_P.value < 0.1 ~ "Heterogeneous"
    )
  )

meta_analysis_subset_with_PD_metaAnalysis <- full_join(constipation_joined_analyses_sub, meta_analysis_subset, by = "FeatureID") %>% 
  relocate(PD_assoc_metaAnalysis, .after = 'FeatureID') %>% 
  mutate(PD_assoc_metaAnalysis = ifelse(is.na(PD_assoc_metaAnalysis), "Not tested", PD_assoc_metaAnalysis))
```

```{r, fig.height=6, fig.width=10}
colors <- c("NS" = "gray", "Increased" = "steelblue3", "Decreased" = "red3", "Not tested" = "white") 

Beta_plot <- meta_analysis_subset_with_PD_metaAnalysis %>% 
ggplot(aes(Beta.Asnicar, Beta.Wallen, fill = PD_assoc_metaAnalysis)) + 
  geom_point(pch = 21, size = 2.5)+
  scale_color_manual(values = colors, aesthetics = c("color", "fill")) + 
  stat_cor(method = "pearson", mapping = aes(group = 1)) +
  geom_abline(intercept = 0, slope = 1, color = "red") + 
  labs(
    fill = "Meta-Analysis (PD/Control) associations"
  )

CohenD_plot <- meta_analysis_subset_with_PD_metaAnalysis %>% 
ggplot(aes(CohenD.Asnicar, CohenD.Wallen, fill = PD_assoc_metaAnalysis)) + 
  geom_point(pch = 21, size = 2.5)+
  scale_color_manual(values = colors, aesthetics = c("color", "fill")) + 
  stat_cor(method = "pearson", mapping = aes(group = 1)) +
  geom_abline(intercept = 0, slope = 1, color = "red") + 
  labs(
    fill = "Meta-Analysis (PD/Control) associations"
  )

ggarrange(Beta_plot, CohenD_plot, common.legend = TRUE)
```

# Session Info

```{r}
sessionInfo()
```
